---
title: "treesort reassortment inference"
author: Elgin Akin 
date: 2025-05-15
format: html
---

# OPtions for reassortment inference 

- treesort
- treeknit
- colre

# TreeSort 
- re-download fasta files for treesort - include a different name schema than for nextstrain builds `sample_ID|date`. modify fludb accession download scripts to incorperate segments and headers 


## Rename Treetips to include dates

```{shell}
python scripts/date_tree_tips.py \
    -t results/vic/ha/tree_raw.nwk \
    -m results/vic/ha/filtered.tsv \
    -f results/vic/ha/aligned.fasta \
    -O TreeSort/data/ha_tree_raw.aln \
    --strain-column "sample_ID" \
    --date-column "date" \
    --verbosity 2 \
    -o TreeSort/data/ha_tree_raw.nwk | tee -a logs/tipdates.out

python scripts/date_tree_tips.py \
    -t results/vic/na/tree_raw.nwk \
    -m results/vic/na/filtered.tsv \
    -f results/vic/na/aligned.fasta \
    -O TreeSort/data/na_tree_raw.aln \
    --strain-column "sample_ID" \
    --date-column "date" \
    --verbosity 2 \
    -o TreeSort/data/na_tree_raw.nwk | tee -a logs/tipdates.out

python scripts/date_tree_tips.py \
    -t results/vic/pb2/tree_raw.nwk \
    -m results/vic/pb2/filtered.tsv \
    -f results/vic/pb2/aligned.fasta \
    -O TreeSort/data/pb2_tree_raw.aln \
    --strain-column "sample_ID" \
    --date-column "date" \
    --verbosity 2 \
    -o TreeSort/data/pb2_tree_raw.nwk | tee -a logs/tipdates.out

python scripts/date_tree_tips.py \
    -t results/vic/pb1/tree_raw.nwk \
    -m results/vic/pb1/filtered.tsv \
    -f results/vic/pb1/aligned.fasta \
    -O TreeSort/data/pb1_tree_raw.aln \
    --strain-column "sample_ID" \
    --date-column "date" \
    --verbosity 2 \
    -o TreeSort/data/pb1_tree_raw.nwk | tee -a logs/tipdates.out

python scripts/date_tree_tips.py \
    -t results/vic/pa/tree_raw.nwk \
    -m results/vic/pa/filtered.tsv \
    -f results/vic/pa/aligned.fasta \
    -O TreeSort/data/pa_tree_raw.aln \
    --strain-column "sample_ID" \
    --date-column "date" \
    --verbosity 2 \
    -o TreeSort/data/pa_tree_raw.nwk | tee -a logs/tipdates.out

python scripts/date_tree_tips.py \
    -t results/vic/np/tree_raw.nwk \
    -m results/vic/np/filtered.tsv \
    -f results/vic/np/aligned.fasta \
    -O TreeSort/data/np_tree_raw.aln \
    --strain-column "sample_ID" \
    --date-column "date" \
    --verbosity 2 \
    -o TreeSort/data/np_tree_raw.nwk | tee -a logs/tipdates.out

python scripts/date_tree_tips.py \
    -t results/vic/mp/tree_raw.nwk \
    -m results/vic/mp/filtered.tsv \
    -f results/vic/mp/aligned.fasta \
    -O TreeSort/data/mp_tree_raw.aln \
    --strain-column "sample_ID" \
    --date-column "date" \
    --verbosity 2 \
    -o TreeSort/data/mp_tree_raw.nwk | tee -a logs/tipdates.out

python scripts/date_tree_tips.py \
    -t results/vic/ns/tree_raw.nwk \
    -m results/vic/ns/filtered.tsv \
    -f results/vic/ns/aligned.fasta \
    -O TreeSort/data/ns_tree_raw.aln \
    --strain-column "sample_ID" \
    --date-column "date" \
    --verbosity 2 \
    -o TreeSort/data/ns_tree_raw.nwk | tee -a logs/tipdates.out

```


## generate descriptor file

Done manually

## root ha tree ?
yes required 
```{shell}
python TreeSort/treetime-root.py TreeSort/data/ha_tree_raw.nwk TreeSort/data/ha_tree_raw.aln
python TreeSort/treetime-root.py TreeSort/data/na_tree_raw.nwk TreeSort/data/na_tree_raw.aln
python TreeSort/treetime-root.py TreeSort/data/pb2_tree_raw.nwk TreeSort/data/pb2_tree_raw.aln
python TreeSort/treetime-root.py TreeSort/data/pb1_tree_raw.nwk TreeSort/data/pb1_tree_raw.aln
python TreeSort/treetime-root.py TreeSort/data/pa_tree_raw.nwk TreeSort/data/pa_tree_raw.aln
python TreeSort/treetime-root.py TreeSort/data/np_tree_raw.nwk TreeSort/data/np_tree_raw.aln
python TreeSort/treetime-root.py TreeSort/data/mp_tree_raw.nwk TreeSort/data/mp_tree_raw.aln
python TreeSort/treetime-root.py TreeSort/data/ns_tree_raw.nwk TreeSort/data/ns_tree_raw.aln
```

## check tree and fasta congruency

```{shell}
python scripts/check_tree_fasta_congruency.py \
  -t TreeSort/data/ha_tree_raw.nwk \
  -f TreeSort/data/ha_tree_raw.aln \
  -o TreeSort/data/ha_tree_fasta_congruency.tsv
```

## strain names need to have dates within the name...

```{shell}
treesort \
    -i TreeSort/data/descriptor_hana.csv \
    -o TreeSort/results/ibv_hana_annotated.tre \
    --clades TreeSort/results/ibv_hana_clades.txt \
    --pvalue
```

```{shell}
treesort \
    -i TreeSort/data/descriptor.csv \
    -o results/treesort/ibv_ha_annotated.tre \
    --clades results/treesort/clades.txt
```

## okay lets try to use the augur refined tree (should be rooted) as input.
Not able to find the reassortment event...
```{shell}
python scripts/date_tree_tips.py \
    -t results/vic/ha/tree.nwk \
    -m results/vic/ha/filtered.tsv \
    --strain-column "sample_ID" \
    --date-column "date" \
    --verbosity 2 \
    -o TreeSort/data/ha_tree.nwk | tee -a logs/tipdates.out

python scripts/date_tree_tips.py \
    -t results/vic/na/tree.nwk \
    -m results/vic/na/filtered.tsv \
    --strain-column "sample_ID" \
    --date-column "date" \
    --verbosity 2 \
    -o TreeSort/data/na_tree.nwk | tee -a logs/tipdates.out

python scripts/date_tree_tips.py \
    -t results/vic/pb2/tree.nwk \
    -m results/vic/pb2/filtered.tsv \
    --strain-column "sample_ID" \
    --date-column "date" \
    --verbosity 2 \
    -o TreeSort/data/pb2_tree.nwk | tee -a logs/tipdates.out

python scripts/date_tree_tips.py \
    -t results/vic/pb1/tree.nwk \
    -m results/vic/pb1/filtered.tsv \
    --strain-column "sample_ID" \
    --date-column "date" \
    --verbosity 2 \
    -o TreeSort/data/pb1_tree.nwk | tee -a logs/tipdates.out

python scripts/date_tree_tips.py \
    -t results/vic/pa/tree.nwk \
    -m results/vic/pa/filtered.tsv \
    --strain-column "sample_ID" \
    --date-column "date" \
    --verbosity 2 \
    -o TreeSort/data/pa_tree.nwk | tee -a logs/tipdates.out

python scripts/date_tree_tips.py \
    -t results/vic/np/tree.nwk \
    -m results/vic/np/filtered.tsv \
    --strain-column "sample_ID" \
    --date-column "date" \
    --verbosity 2 \
    -o TreeSort/data/np_tree.nwk | tee -a logs/tipdates.out

python scripts/date_tree_tips.py \
    -t results/vic/mp/tree.nwk \
    -m results/vic/mp/filtered.tsv \
    --strain-column "sample_ID" \
    --date-column "date" \
    --verbosity 2 \
    -o TreeSort/data/mp_tree.nwk | tee -a logs/tipdates.out

python scripts/date_tree_tips.py \
    -t results/vic/ns/tree.nwk \
    -m results/vic/ns/filtered.tsv \
    --strain-column "sample_ID" \
    --date-column "date" \
    --verbosity 2 \
    -o TreeSort/data/ns_tree.nwk | tee -a logs/tipdates.out
```

update descriptor with augur refine renamed tip trees 
```{shell}
treesort \
    -i TreeSort/data/descriptor.csv \
    -o TreeSort/results/ibv_hana_annotated.tre \
    --clades TreeSort/results/ibv_hana_clades.txt
```

## try adjusting p val and switch back to the non-refined branch trees

```{shell}
treesort \
    -i TreeSort/data/descriptor_hana.csv \
    -o TreeSort/results/ibv_testpval_annotated.tre \
    --pvalue 0.01 \
    --clades TreeSort/results/ibv_testpval.clades
```



# Attempting CoalRe 

https://taming-the-beast.org/tutorials/Reassortment-Tutorial/#:~:text=Download%20CoalRe,and%20download%20the%20CoalRe%20package%20.&text=After%20the%20package%20is%20installed%2C%20re%2Dstart%20BEAUti.

# Estimating Nucleotide Diversities from Alignments 

Its possible that the slow mutation rate of the NS segment has complicated ancestry prediction. 
- Calulate pi for each segment alignment 
- calculate root-to-tip regression for each segment and plot...


## root-to-tip regression


Root to tip regressions were calculated using treetime called through TreeSort's included [treetime-root.py](TreeSort/treetime-root.py)

### Plot calcualted rtts for 8 segments
```{r}
plot_rtt_regression <- function(data, gene_name) {
  # Fit linear model
  model <- lm(`root-to-tip distance` ~ date, data = data)
  
  # Extract values
  intercept <- coef(model)[1]
  slope <- coef(model)[2]
  slope_se <- summary(model)$coefficients["date", "Std. Error"]
  r_squared <- summary(model)$r.squared
  root_date <- -intercept / slope

  # 95% CI
  ci <- confint(model)["date", ]
  slope_ci_str <- paste0(signif(slope, 2))
  #slope_ci_str <- paste0(signif(slope, 2), " [", signif(ci[1], 2), ", ", signif(ci[2], 2), "]")
  
  # Generate plot
  p <- ggplot(data, aes(x = date, y = `root-to-tip distance`)) +
    geom_point(alpha = 0.6, size = 1.8) +
    geom_smooth(method = "lm", se = TRUE, color = '#2C3E50') +
    annotate("text",
         x = max(data$date),
         y = min(data$`root-to-tip distance`),
         hjust = 1,
         label = paste0("y = α + βt\n",
                        "β = ", slope_ci_str, "\n",
                        "R² = ", signif(r_squared, 3), "\n",
                        "root date: ", round(root_date, 2)),
         size = 4) +
    labs(title = paste("Root-to-tip regression:", gene_name),
         x = "Date",
         y = "Root-to-tip distance") +
    theme_minimal()
  
  return(p)
}


# Extract slope and CI from an RTT dataset
extract_slope_ci <- function(data, gene_name) {
  model <- lm(`root-to-tip distance` ~ date, data = data)
  slope <- coef(model)["date"]
  ci <- confint(model)["date", ]
  tibble(
    gene = gene_name,
    slope = slope,
    ci_lower = ci[1],
    ci_upper = ci[2]
  )
}

# Plot bar plot of slopes with error bars
plot_all_slopes <- function(results_df) {
  results_df <- results_df |> 
    arrange(desc(slope)) |> 
    mutate(gene = factor(gene, levels = gene))  # Preserve order in plot

  ggplot(results_df, aes(x = gene, y = slope)) +
    geom_col(fill = "steelblue", alpha = 0.8) +
    geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2) +
    labs(
      title = "Estimated substitution rate (β) with 95% CI",
      x = "Segment",
      y = "Slope β"
    ) +
    theme_minimal(base_size = 14)
}

```


```{r}

library(tidyverse)

rtt_pb2 = read_csv("Treesort/data/treetime_root/pb2_tree_raw.aln.treetime/rtt.csv", skip = 1) |> 
    filter(!str_detect(name, "NODE"))
rtt_pb1 = read_csv("Treesort/data/treetime_root/pb1_tree_raw.aln.treetime/rtt.csv", skip = 1) |> 
    filter(!str_detect(name, "NODE"))
rtt_pa = read_csv("Treesort/data/treetime_root/pa_tree_raw.aln.treetime/rtt.csv", skip = 1) |> 
    filter(!str_detect(name, "NODE"))
rtt_ha = read_csv("Treesort/data/treetime_root/ha_tree_raw.aln.treetime/rtt.csv", skip = 1) |> 
    filter(!str_detect(name, "NODE"))
rtt_np = read_csv("Treesort/data/treetime_root/np_tree_raw.aln.treetime/rtt.csv", skip = 1) |> 
    filter(!str_detect(name, "NODE"))
rtt_na = read_csv("Treesort/data/treetime_root/na_tree_raw.aln.treetime/rtt.csv", skip = 1) |> 
    filter(!str_detect(name, "NODE"))
rtt_mp = read_csv("Treesort/data/treetime_root/mp_tree_raw.aln.treetime/rtt.csv", skip = 1) |> 
    filter(!str_detect(name, "NODE"))
rtt_ns = read_csv("Treesort/data/treetime_root/ns_tree_raw.aln.treetime/rtt.csv", skip = 1) |> 
    filter(!str_detect(name, "NODE"))

p_pb2 <- plot_rtt_regression(rtt_pb2, "PB2")
p_pb1 <- plot_rtt_regression(rtt_pb1, "PB1")
p_pa  <- plot_rtt_regression(rtt_pa,  "PA")
p_ha  <- plot_rtt_regression(rtt_ha,  "HA")
p_np  <- plot_rtt_regression(rtt_np,  "NP")
p_na  <- plot_rtt_regression(rtt_np,  "NA")
p_mp  <- plot_rtt_regression(rtt_na,  "MP")
p_ns  <- plot_rtt_regression(rtt_ns,  "NS")

library(patchwork)

(p_ha | p_na) /
(p_pb2  | p_pb1  | p_pa) /
(p_np | p_mp | p_ns)

```





```{r}
# Helper to read RTT files
read_rtt_data <- function(path) {
  read_csv(path, skip = 1) |> filter(!str_detect(name, "NODE"))
}

# Collect results
slope_results <- bind_rows(
  extract_slope_ci(read_rtt_data("Treesort/data/treetime_root/pb2_tree_raw.aln.treetime/rtt.csv"), "PB2"),
  extract_slope_ci(read_rtt_data("Treesort/data/treetime_root/pb1_tree_raw.aln.treetime/rtt.csv"), "PB1"),
  extract_slope_ci(read_rtt_data("Treesort/data/treetime_root/pa_tree_raw.aln.treetime/rtt.csv"),  "PA"),
  extract_slope_ci(read_rtt_data("Treesort/data/treetime_root/ha_tree_raw.aln.treetime/rtt.csv"),  "HA"),
  extract_slope_ci(read_rtt_data("Treesort/data/treetime_root/np_tree_raw.aln.treetime/rtt.csv"),  "NP"),
  extract_slope_ci(read_rtt_data("Treesort/data/treetime_root/na_tree_raw.aln.treetime/rtt.csv"),  "NA"),
  extract_slope_ci(read_rtt_data("Treesort/data/treetime_root/ns_tree_raw.aln.treetime/rtt.csv"),  "NS")
)

# Plot summary of all slopes
plot_all_slopes(slope_results)

```


# try fancier

```{r}
plot_rtt_regression <- function(data, gene_name, label_position = "top-left") {
  # Fit linear model
  model <- lm(`root-to-tip distance` ~ date, data = data)
  
  # Extract values
  intercept <- coef(model)[1]
  slope <- coef(model)[2]
  slope_se <- summary(model)$coefficients["date", "Std. Error"]
  r_squared <- summary(model)$r.squared
  root_date <- -intercept / slope
  ci <- confint(model)["date", ]
  slope_ci_str <- paste0(signif(slope, 2))

  # Create label text
  label_text <- paste0("y = α + βt\n",
                       "β = ", slope_ci_str, "\n",
                       "R² = ", signif(r_squared, 3), "\n",
                       "root date: ", round(root_date, 2))

  # Define label position
  x_pos <- if (label_position == "top-left") min(data$date) else max(data$date)
  y_pos <- if (label_position == "top-left") max(data$`root-to-tip distance`) else min(data$`root-to-tip distance`)
  h_just <- if (label_position == "top-left") 0 else 1

  # Create plot
  p <- ggplot(data, aes(x = date, y = `root-to-tip distance`)) +
    geom_point(alpha = 0.6, size = 1.8) +
    geom_smooth(method = "lm", se = TRUE, color = '#2C3E50') +
    geom_label(
      aes(x = x_pos, y = y_pos, label = label_text),
      hjust = h_just,
      vjust = if (label_position == "top-left") 1 else 0,
      fill = alpha("white", 0.75),
      color = "black",
      label.size = 0.4,
      size = 4,
      inherit.aes = FALSE
    ) +
    labs(
      title = paste("Root-to-tip regression:", gene_name),
      x = "Date",
      y = "Root-to-tip distance"
    ) +
    theme_minimal()
  
  return(p)
}


plot_rtt_regression(rtt_pb2, "HA", label_position = "top-left")

```

## pi 

```

```

