---
title: "Observational Case Frequencies"
author: Elgin Akin
date: 2025-06-15
format: html
---


- use the `source/intermediate/vic/ha/metadata_qc_location_clades_dedup_clades.tsv` as the input data frame
- de-duplicated by strain name
- location column expanded to region / country / division / location
    - will need to look at the number of missing rows within each category...


# Figures 

## Total Frequency Summarized by Month
1. Global Frequency by Subclade
2. North American Frequency by subclade
3. Baltimore/JHH Frequency by Subclade
4. D197E Glycosylation by Month 

## Maps Global Distribution 

1. C.3 distribution globally
2. C.5.1 distribution globally
3. C.5.6 distribution globally
4. C.5.7 distribution globally


# Config

```{r}
subclade_colors <- c("C.5.1" = "#d95f02", 
                     "C.5.5" = "#66c2a5", 
                     "C.5.6" = "#051334",
                     "C.5.7" = "#666666",
                     "C.5" = "#e6ab02", 
                     "C.3" = "#00A7E1")

subclade_colors <- c("C.5.7" = "#666666",
                      "C.5.6" = "#051334",
                      "C.5.5" = "#66c2a5",
                      "C.5.4" = "#575482",
                      "C.5.3" = "#7570b3",
                      "C.5.2" = "#9893E1",
                      "C.5.1" = "#d95f02",
                      "C.5" = "#623C74", 
                      "C.4" = "#E4B19B",
                      "C.3" = "#00A7E1",
                      "C.2" = "#F2CACA",
                      "C.1" = "#F8CFCF",
                      "C" = "#FFE2D8",
                      "A" = "#F9DC5C",
                      "A.3.2" = "#FAE588",
                      "A.3.1.1" = "#FCEFB4",
                      "A.3" = "#FDF4CB")

#Other pallettes: 
 # - https://loading.io/color/feature/DragonBall-%E3%83%89%E3%83%A9%E3%82%B4%E3%83%B3%E3%83%9C%E3%83%BC%E3%83%AB

#Dark2
#1b9e77
#d95f02
#7570b3
#e7298a
#66a61e
#e6ab02
#a6761d
#666666

```


## Case Counts

```{r}
# Load necessary libraries
library(tidyverse)
library(ggplot2)
library(tidyr)
library(dplyr)
library(lubridate)

flu_data <- read_tsv("data/case_counts_2425.txt")

# Convert date column to Date type
flu_data$date_range <- as.Date(flu_data$date_range, format = "%Y-%m-%d")


# Reshape data to long format
flu_long <- flu_data %>%
  pivot_longer(cols = c(IAV_clinic, IBV_clinic), names_to = "Virus", values_to = "Case Count") %>%
  mutate(Virus = recode(Virus,
                        "IAV_clinic" = "Influenza A",
                        "IBV_clinic" = "Influenza B"))

                
# Filter to only show dates between October 2023 and April 2024
filtered_flu <- flu_long %>% filter(date_range >= as.Date("2024-10-02") & date_range <= as.Date("2025-12-31"))

# fill 2019-2024 df 
# summarized
filtered_fliu_monthly_dates <- seq(floor_date(min(filtered_flu$date_range), "month"),
                     floor_date(max(filtered_flu$date_range), "month"),
                     by = "month")

case_count = filtered_flu |>
  ggplot(aes(date_range, `Case Count`, color = Virus)) + 
  geom_smooth(se = FALSE, linewidth = 1.2) + 
  scale_x_date(date_labels = "%b\n%y",date_breaks = "1 month", expand = expansion(mult = c(0.05, 0.1))) +
  scale_color_manual(values = c("Influenza A" = "black", "Influenza B" = "red")) +
  labs(y = "Case Count", x = "Case Date", color = "Virus") +
  theme_classic(base_size = 16, base_family = "Helvetica") +
  theme(
    legend.position = "top",
    axis.title = element_text(size = 18, face = "bold"),
    axis.text = element_text(size = 16),
    legend.text = element_text(size = 15),
    legend.title = element_text(size = 15)
  )

case_count

ggsave('figures/case_count.png', case_count, height = 6, width = 6)
```

## Total Frequency Summarized by Month

```{r}
library(tidyverse)
library(ggstream)
library(plotly)
library(patchwork)

# import nextclade data 
nextclade <- read_tsv("source/intermediate/vic/ha/metadata_qc_location_clades_dedup_clades.tsv") |>
  mutate(date = if_else(str_detect(date, "2025-..-..") & !str_detect(date, "2025-\\d{2}-\\d{2}"),
                        "2025-01-01", date),
         date = as.Date(date))

#filter out vaccine strain
nextclade_filt = nextclade |> 
  filter(date >= as.Date("2022-01-01") & date <= as.Date("2025-12-31")) |>
  filter(database_origin == "mostafa_lab") 
  
# Count occurrences of each subclade per date
frequency <-  nextclade_filt |> 
  count(date, subclade) |>
  ungroup()

# Generate sequence of dates representing each week
weekly_dates <- seq(min(frequency$date), max(frequency$date), by = "week")

monthly_dates <- seq(floor_date(min(frequency$date), "month"),
                     floor_date(max(frequency$date), "month"),
                     by = "month")

# Create the base plot
p1 = frequency |> 
  ggplot(aes(date, n, fill=subclade), color = "black") + 
  scale_fill_manual(values = c(subclade_colors, "gray")) +
  theme_classic()

ridge = p1 +
  geom_stream(sorting = "inside_out", bw = 1.2, type = "proportional", color = "black",) +
  #geom_vline(xintercept = as.numeric(monthly_dates), linetype = "solid", color = "black", alpha = 0.1) +
  scale_x_date(date_labels = "%b \n %y", expand = expansion(mult = c(0.05, 0.1))) +  #
  theme(legend.position = "top",
        axis.title = element_text(size = 18, face = "bold"),
        axis.text = element_text(size = 16),
        legend.text = element_text(size = 15),  # Increase legend text size
        legend.title = element_text(size = 15)) +  # Increase legend title size
  labs(y="Proportion", x="Sequencing Date")

ggsave('figures/jhh_frequency_proportional.png', ridge, height = 6, width = 6)
ggsave('figures/jhh_supplemental_frequency_proportiona.png', ridge, height = 4, width = 8)

ridge
```

### filter only to north america

```{r}
library(tidyverse)
library(ggstream)
library(plotly)
library(patchwork)


# import nextclade data 
nextclade <- read_tsv("source/intermediate/vic/ha/metadata_qc_location_clades_dedup_clades.tsv") |>
  mutate(date = if_else(str_detect(date, "2025-..-..") & !str_detect(date, "2025-\\d{2}-\\d{2}"),
                        "2025-01-01", date),
         date = as.Date(date))

#filter out vaccine strain
nextclade_filt <- nextclade |> 
  filter(
    date >= as.Date("2023-10-01") & date <= as.Date("2025-12-31"),
    database_origin != "mostafa_lab",
    country == "United States"
  )
  
# Count occurrences of each subclade per date
frequency <-  nextclade_filt |> 
  count(date, subclade) |>
  ungroup()

frequency <- nextclade_filt |>
  filter(!is.na(subclade), !is.na(date)) |>
  mutate(month = floor_date(date, unit = "month")) |>
  count(month, subclade) |>
  rename(date = month) |>
  ungroup()

# Generate sequence of dates representing each week
weekly_dates <- seq(min(frequency$date), max(frequency$date), by = "week")

monthly_dates <- seq(floor_date(min(frequency$date), "month"),
                     floor_date(max(frequency$date), "month"),
                     by = "month")

# Create the base plot
p1 = frequency |> 
  ggplot(aes(date, n, fill=subclade), color = "black") + 
  scale_fill_manual(values = c(subclade_colors, "gray")) +
  theme_classic()

ridge = p1 +
  geom_stream(sorting = "inside_out", bw = 0.7, type = "proportional", color = "black",) +
  #geom_vline(xintercept = as.numeric(monthly_dates), linetype = "solid", color = "black", alpha = 0.1) +
  scale_x_date(date_labels = "%b \n %y", expand = expansion(mult = c(0.05, 0.1))) +  #
  theme(legend.position = "top",
        axis.title = element_text(size = 18, face = "bold"),
        axis.text = element_text(size = 16),
        legend.text = element_text(size = 15),  # Increase legend text size
        legend.title = element_text(size = 15)) +  # Increase legend title size
  labs(y="Proportion", x="Sequencing Date")

ggsave('figures/NHfrequency_ridgeline.png', ridge, height = 4, width = 8)

ridge
```


## Global 
```{r}
library(tidyverse)
library(ggstream)
library(plotly)
library(patchwork)

# import nextclade data 
nextclade <- read_tsv("source/intermediate/vic/ha/metadata_qc_location_clades_dedup_clades.tsv") |>
  mutate(date = if_else(str_detect(date, "2025-..-..") & !str_detect(date, "2025-\\d{2}-\\d{2}"),
                        "2025-01-01", date),
         date = as.Date(date))

#filter out vaccine strain
nextclade_filt <- nextclade |> 
  filter(
    date >= as.Date("2023-10-01") & date <= as.Date("2025-12-31"),
    database_origin != "mostafa_lab"
  )
  
# Count occurrences of each subclade per date
frequency <-  nextclade_filt |> 
  count(date, subclade) |>
  ungroup()

frequency <- nextclade_filt |>
  filter(!is.na(subclade), !is.na(date)) |>
  mutate(month = floor_date(date, unit = "month")) |>
  count(month, subclade) |>
  rename(date = month) |>
  ungroup()

# Generate sequence of dates representing each week
weekly_dates <- seq(min(frequency$date), max(frequency$date), by = "week")

monthly_dates <- seq(floor_date(min(frequency$date), "month"),
                     floor_date(max(frequency$date), "month"),
                     by = "month")

# Create the base plot
p1 = frequency |> 
  ggplot(aes(date, n, fill=subclade), color = "black") + 
  scale_fill_manual(values = c(subclade_colors, "gray")) +
  theme_classic()

ridge = p1 +
  geom_stream(sorting = "inside_out", bw = 0.7, type = "proportional", color = "black",) +
  #geom_vline(xintercept = as.numeric(monthly_dates), linetype = "solid", color = "black", alpha = 0.1) +
  scale_x_date(date_labels = "%b \n %y", expand = expansion(mult = c(0.05, 0.1))) +  #
  theme(legend.position = "top",
        axis.title = element_text(size = 18, face = "bold"),
        axis.text = element_text(size = 16),
        legend.text = element_text(size = 15),  # Increase legend text size
        legend.title = element_text(size = 15)) +  # Increase legend title size
  labs(y="Proportion", x="Sequencing Date")

ggsave('figures/global_frequency_ridgeline.png', ridge, height = 4, width = 8)

ridge
```


# Aggregated subclade table

```{r}
library(tidyverse)
library(gt)

nextclade_filt_table <- nextclade |> 
  filter(
    date >= as.Date("2024-10-01") & date <= as.Date("2025-12-31")
  )

# Filter and summarize
subclade_summary <- nextclade_filt_table |> 
  filter(!is.na(subclade)) |> 
  mutate(location_group = case_when(
    division == "Maryland" ~ "Maryland",
    country == "United States" ~ "United States",
    TRUE ~ NA_character_
  )) |> 
  filter(!is.na(location_group)) |> 
  count(location_group, subclade) |> 
  pivot_wider(names_from = location_group, values_from = n, values_fill = 0) |> 
  mutate(
    `Maryland Percent` = Maryland / sum(Maryland) * 100,
    `United States Percent` = `United States` / sum(`United States`) * 100
  ) |> 
  arrange(desc(`United States`)) |> 
  rename(
    `Subclade` = subclade,
    `Maryland Count` = Maryland,
    `United States Count` = `United States`
  )

# Format with gt
subclade_summary |> 
  gt() |> 
  fmt_number(columns = c(`Maryland Percent`, `United States Percent`), decimals = 1) |> 
  cols_label(
    `Subclade` = "Subclade",
    `Maryland Count` = "C Maryland Count",
    `Maryland Percent` = "Maryland Percent",
    `United States Count` = "North America Count",
    `United States Percent` = "North America Percent"
  ) |> 
  tab_header(
    title = "IBV Subclade Sequence Submissions\nin Maryland vs North America"
  )

```