---
title: "Observational Case Frequencies"
author: Elgin Akin
date: 2025-06-15
format: html
---

-   use the `source/intermediate/vic/ha/metadata_qc_location_clades_dedup_clades_glycans.tsv` as the input data frame
-   de-duplicated by strain name
-   location column expanded to region / country / division / location
    -   will need to look at the number of missing rows within each category...

# Figures

## Total Frequency Summarized by Month

1.  Global Frequency by Subclade
2.  North American Frequency by subclade
3.  Baltimore/JHH Frequency by Subclade
4.  D197E Glycosylation by Month

## Maps Global Distribution

1.  C.3 distribution globally
2.  C.5.1 distribution globally
3.  C.5.6 distribution globally
4.  C.5.7 distribution globally

# Config

```{r}
subclade_colors <- c("C.5.1" = "#d95f02", 
                     "C.5.5" = "#66c2a5", 
                     "C.5.6" = "#051334",
                     "C.5.7" = "#666666",
                     "C.5" = "#e6ab02", 
                     "C.3" = "#00A7E1")

subclade_colors <- c("C.5.7" = "#666666",
                      "C.5.6" = "#051334",
                      "C.5.5" = "#66c2a5",
                      "C.5.4" = "#575482",
                      "C.5.3" = "#7570b3",
                      "C.5.2" = "#9893E1",
                      "C.5.1" = "#d95f02",
                      "C.5" = "#623C74", 
                      "C.4" = "#E4B19B",
                      "C.3" = "#00A7E1",
                      "C.2" = "#F2CACA",
                      "C.1" = "#F8CFCF",
                      "C" = "#FFE2D8",
                      "A" = "#F9DC5C",
                      "A.3.2" = "#FAE588",
                      "A.3.1.1" = "#FCEFB4",
                      "A.3" = "#FDF4CB")

#Other pallettes: 
 # - https://loading.io/color/feature/DragonBall-%E3%83%89%E3%83%A9%E3%82%B4%E3%83%B3%E3%83%9C%E3%83%BC%E3%83%AB

#Dark2
#1b9e77
#d95f02
#7570b3
#e7298a
#66a61e
#e6ab02
#a6761d
#666666

```

# Case Counts

```{r}
# Load necessary libraries
library(tidyverse)
library(ggplot2)
library(tidyr)
library(dplyr)
library(lubridate)

flu_data <- read_tsv("data/case_counts_2425.txt")

# Convert date column to Date type
flu_data$date_range <- as.Date(flu_data$date_range, format = "%Y-%m-%d")


# Reshape data to long format
flu_long <- flu_data %>%
  pivot_longer(cols = c(IAV_clinic, IBV_clinic), names_to = "Virus", values_to = "Case Count") %>%
  mutate(Virus = recode(Virus,
                        "IAV_clinic" = "Influenza A",
                        "IBV_clinic" = "Influenza B"))

                
# Filter to only show dates between October 2023 and April 2024
filtered_flu <- flu_long %>% filter(date_range >= as.Date("2024-10-02") & date_range <= as.Date("2025-12-31"))

# fill 2019-2024 df 
# summarized
filtered_fliu_monthly_dates <- seq(floor_date(min(filtered_flu$date_range), "month"),
                     floor_date(max(filtered_flu$date_range), "month"),
                     by = "month")

case_count = filtered_flu |>
  ggplot(aes(date_range, `Case Count`, color = Virus)) + 
  geom_smooth(se = FALSE, linewidth = 1.2) + 
  scale_x_date(date_labels = "%b\n%y",date_breaks = "1 month", expand = expansion(mult = c(0.05, 0.1))) +
  scale_color_manual(values = c("Influenza A" = "black", "Influenza B" = "red")) +
  labs(y = "Case Count", x = "Case Date", color = "Virus") +
  theme_classic(base_size = 16, base_family = "Helvetica") +
  theme(
    legend.position = "top",
    axis.title = element_text(size = 18, face = "bold"),
    axis.text = element_text(size = 16),
    legend.text = element_text(size = 15),
    legend.title = element_text(size = 15)
  )

case_count

ggsave('figures/case_count.png', case_count, height = 5, width = 5)
```


# Table of clinical cases by Month
```{r}
library(tidyverse)
library(gt)

# Raw data
clinic_counts <- tibble(
  date_range = as.Date(c(
    "2024-10-01", "2024-11-01", "2024-12-01",
    "2025-01-01", "2025-02-01", "2025-03-01", "2025-04-01"
  )),
  IAV_clinic = c(2, 57, 510, 1988, 2229, 526, 53),
  IBV_clinic = c(0, 4, 12, 41, 65, 372, 400)
) |>
  mutate(
    Total = IAV_clinic + IBV_clinic,
    `IAV Percent` = IAV_clinic / Total * 100,
    `IBV Percent` = IBV_clinic / Total * 100
  ) |>
  rename(
    `Date` = date_range,
    `IAV Count` = IAV_clinic,
    `IBV Count` = IBV_clinic
  )

# Add grand total row
clinic_counts_totals <- clinic_counts |>
  summarise(
    Date = as.Date(NA),
    `IAV Count` = sum(`IAV Count`),
    `IBV Count` = sum(`IBV Count`),
    Total = sum(Total),
    `IAV Percent` = `IAV Count` / Total * 100,
    `IBV Percent` = `IBV Count` / Total * 100
  ) |>
  mutate(Date = as.Date("9999-12-31"))  # use dummy future date for sorting

# Bind totals row
final_table <- bind_rows(clinic_counts, clinic_counts_totals) |>
  arrange(Date) |>
  mutate(Date = if_else(Date == as.Date("9999-12-31"), as.Date(NA), Date))

# Format with gt
table_clinical_frequency = final_table |>
                                gt() |>
                                fmt_number(columns = c(`IAV Percent`, `IBV Percent`), decimals = 1) |>
                                #fmt_date(columns = "Date", date_style = "long") |>
                                fmt_date(columns = "Date", date_style = "yMMM") |>
                                cols_label(
                                  Date = "Date",
                                  `IAV Count` = "IAV Count",
                                  `IBV Count` = "IBV Count",
                                  Total = "Total Cases",
                                  `IAV Percent` = "IAV %",
                                  `IBV Percent` = "IBV %"
                                ) |>
                                tab_header(
                                  title = "Clinic-Detected Influenza Cases by Month",
                                  subtitle = "With Total IAV and IBV Counts"
                                ) |>
                                tab_style(
                                  style = cell_text(weight = "bold"),
                                  locations = cells_body(
                                    rows = is.na(Date)
                                  )
                                ) |>
                                text_transform(
                                  locations = cells_body(columns = "Date", rows = is.na(Date)),
                                  fn = function(x) "TOTAL"
                                )

table_clinical_frequency |> 
  gtsave("figures/table_clinic_frequency.docx")

table_clinical_frequency |> gtsave("figures/table_clinic_frequency.pdf")

```

## Total Subclade Frequency Summarized by Month in JHH

```{r}
library(tidyverse)
library(ggstream)
library(plotly)
library(patchwork)

# import nextclade data 
nextclade <- read_tsv("source/intermediate/vic/ha/metadata_qc_location_clades_dedup_clades_glycans.tsv") |>
  mutate(date = if_else(str_detect(date, "2025-..-..") & !str_detect(date, "2025-\\d{2}-\\d{2}"),
                        "2025-01-01", date),
         date = as.Date(date))

#filter out vaccine strain
nextclade_filt = nextclade |> 
  filter(date >= as.Date("2022-01-01") & date <= as.Date("2025-12-31")) |>
  filter(database_origin == "mostafa_lab")

# Count occurrences of each subclade per date
frequency <-  nextclade_filt |> 
  count(date, subclade) |>
  ungroup()

# Generate sequence of dates representing each week
weekly_dates <- seq(min(frequency$date), max(frequency$date), by = "week")

monthly_dates <- seq(floor_date(min(frequency$date), "month"),
                     floor_date(max(frequency$date), "month"),
                     by = "month")

# Create the base plot for subclade frequency
p1 = frequency |> 
  ggplot(aes(date, n, fill=subclade), color = "black") + 
  scale_fill_manual(values = c(subclade_colors, "gray")) +
  theme_classic()

ridge = p1 +
  geom_stream(sorting = "inside_out", bw = 1.2, type = "ridge", color = "black",) +
  #geom_vline(xintercept = as.numeric(monthly_dates), linetype = "solid", color = "black", alpha = 0.1) +
  scale_x_date(date_labels = "%b \n %y", expand = expansion(mult = c(0.05, 0.1))) +
  scale_y_continuous(breaks = c(0, 50, 100, 150, 200)) +
  theme(legend.position = "none",
        axis.title = element_text(size = 18, face = "bold"),
        axis.text = element_text(size = 16),
        legend.text = element_text(size = 15),  # Increase legend text size
        legend.title = element_text(size = 15)) +  # Increase legend title size
  labs(y="Sequence count", x="Sequencing Date")
proportional = p1 +
  geom_stream(sorting = "inside_out", bw = 1.2, type = "proportional", color = "black",) +
  #geom_vline(xintercept = as.numeric(monthly_dates), linetype = "solid", color = "black", alpha = 0.1) +
  scale_x_date(date_labels = "%b \n %y", expand = expansion(mult = c(0.05, 0.1))) +  #
  theme(legend.position = "top",
        axis.title = element_text(size = 18, face = "bold"),
        axis.text = element_text(size = 16),
        legend.text = element_text(size = 15),  # Increase legend text size
        legend.title = element_text(size = 15)) +  # Increase legend title size
  labs(y="Proportion", x="Sequencing Date")


ggsave('figures/jhh_frequency_ridge.png', ridge, height = 5, width = 5)
ggsave('figures/jhh_supplemental_frequency_ridge.png', ridge, height = 4, width = 8)
ggsave('figures/jhh_supplemental_frequency_proportional.png', proportional, height = 4, width = 8)

ridge
proportional
```

### filter only to north america
```{r}
library(tidyverse)
library(ggstream)
library(plotly)
library(patchwork)


# import nextclade data 
nextclade <- read_tsv("source/intermediate/vic/ha/metadata_qc_location_clades_dedup_clades_glycans.tsv") |>
  mutate(date = if_else(str_detect(date, "2025-..-..") & !str_detect(date, "2025-\\d{2}-\\d{2}"),
                        "2025-01-01", date),
         date = as.Date(date))

#filter out vaccine strain
nextclade_filt <- nextclade |> 
  filter(
    date >= as.Date("2023-10-01") & date <= as.Date("2025-12-31"),
    database_origin != "mostafa_lab",
    country == "United States"
  )
  
# Count occurrences of each subclade per date
frequency <-  nextclade_filt |> 
  count(date, subclade) |>
  ungroup()

frequency <- nextclade_filt |>
  filter(!is.na(subclade), !is.na(date)) |>
  mutate(month = floor_date(date, unit = "month")) |>
  count(month, subclade) |>
  rename(date = month) |>
  ungroup()

# Generate sequence of dates representing each week
weekly_dates <- seq(min(frequency$date), max(frequency$date), by = "week")

monthly_dates <- seq(floor_date(min(frequency$date), "month"),
                     floor_date(max(frequency$date), "month"),
                     by = "month")

# Create the base plot
p1 = frequency |> 
  ggplot(aes(date, n, fill=subclade), color = "black") + 
  scale_fill_manual(values = c(subclade_colors, "gray")) +
  theme_classic()

ridge = p1 +
  geom_stream(sorting = "inside_out", bw = 1.2, type = "ridge", color = "black",) +
  #geom_vline(xintercept = as.numeric(monthly_dates), linetype = "solid", color = "black", alpha = 0.1) +
  scale_x_date(date_labels = "%b \n %y", expand = expansion(mult = c(0.05, 0.1))) +  #
  theme(legend.position = "top",
        axis.title = element_text(size = 18, face = "bold"),
        axis.text = element_text(size = 16),
        legend.text = element_text(size = 15),  # Increase legend text size
        legend.title = element_text(size = 15)) +  # Increase legend title size
  labs(y="Sequence count", x="Sequencing Date")

proportional = p1 +
  geom_stream(sorting = "inside_out", bw = 1.2, type = "proportional", color = "black",) +
  #geom_vline(xintercept = as.numeric(monthly_dates), linetype = "solid", color = "black", alpha = 0.1) +
  scale_x_date(date_labels = "%b \n %y", expand = expansion(mult = c(0.05, 0.1))) +  #
  theme(legend.position = "top",
        axis.title = element_text(size = 18, face = "bold"),
        axis.text = element_text(size = 16),
        legend.text = element_text(size = 15),  # Increase legend text size
        legend.title = element_text(size = 15)) +  # Increase legend title size
  labs(y="Proportion", x="Sequencing Date")

ggsave('figures/NHfrequency_ridgeline.png', ridge, height = 4, width = 8)
ggsave('figures/NHfrequency_proportional.png', proportional, height = 4, width = 8)

ridge
proportional
```

## Global

```{r}
library(tidyverse)
library(ggstream)
library(plotly)
library(patchwork)

# import nextclade data 
nextclade <- read_tsv("source/intermediate/vic/ha/metadata_qc_location_clades_dedup_clades_glycans.tsv") |>
  mutate(date = if_else(str_detect(date, "2025-..-..") & !str_detect(date, "2025-\\d{2}-\\d{2}"),
                        "2025-01-01", date),
         date = as.Date(date))

#filter out vaccine strain
nextclade_filt <- nextclade |> 
  filter(
    date >= as.Date("2023-10-01") & date <= as.Date("2025-12-31"),
    database_origin != "mostafa_lab"
  )
  
# Count occurrences of each subclade per date
frequency <-  nextclade_filt |> 
  count(date, subclade) |>
  ungroup()

frequency <- nextclade_filt |>
  filter(!is.na(subclade), !is.na(date)) |>
  mutate(month = floor_date(date, unit = "month")) |>
  count(month, subclade) |>
  rename(date = month) |>
  ungroup()

# Generate sequence of dates representing each week
weekly_dates <- seq(min(frequency$date), max(frequency$date), by = "week")

monthly_dates <- seq(floor_date(min(frequency$date), "month"),
                     floor_date(max(frequency$date), "month"),
                     by = "month")

# Create the base plot
p1 = frequency |> 
  ggplot(aes(date, n, fill=subclade), color = "black") + 
  scale_fill_manual(values = c(subclade_colors, "gray")) +
  theme_classic()

ridge = p1 +
  geom_stream(sorting = "inside_out", bw = 1.2, type = "ridge", color = "black") +
  #geom_vline(xintercept = as.numeric(monthly_dates), linetype = "solid", color = "black", alpha = 0.1) +
  scale_x_date(date_labels = "%b \n %y", expand = expansion(mult = c(0.05, 0.1))) +  #
  theme(legend.position = "top",
        axis.title = element_text(size = 18, face = "bold"),
        axis.text = element_text(size = 16),
        legend.text = element_text(size = 15),  # Increase legend text size
        legend.title = element_text(size = 15)) +  # Increase legend title size
  labs(y="Sequence count", x="Sequencing Date")

proportional = p1 +
  geom_stream(sorting = "inside_out", bw = 1.2, type = "proportional", color = "black") +
  #geom_vline(xintercept = as.numeric(monthly_dates), linetype = "solid", color = "black", alpha = 0.1) +
  scale_x_date(date_labels = "%b \n %y", expand = expansion(mult = c(0.05, 0.1))) +  #
  theme(legend.position = "top",
        axis.title = element_text(size = 18, face = "bold"),
        axis.text = element_text(size = 16),
        legend.text = element_text(size = 15),  # Increase legend text size
        legend.title = element_text(size = 15)) +  # Increase legend title size
  labs(y="Proportion", x="Sequencing Date")

ggsave('figures/frequency_global_ridgeline.png', ridge, height = 4, width = 8)
ggsave('figures/frequency_global_proportional.png', proportional, height = 4, width = 8)

ridge
proportional
```

# Quantify N-linked glycosylation observation frequency

Use [quantify_hana_glycans.py](scripts/quantify_hana_glycans.py)

```{shell}
python scripts/quantify_hana_glycans.py \
  -i source/intermediate/vic/ha/metadata_qc_location_clades_dedup_clades_glycans.tsv \
  -o source/intermediate/vic/ha/metadata_qc_location_clades_dedup_clades_glycans.tsv \
  -ha ha_glycosylation \
  -na na_glycosylation
```

```{r}
library(tidyverse)
library(lubridate)


glycan_colors <- c(
    "D/EET" = "#72874EFF", 
    "NET" = "#FF746C"
)


# Load data
nextclade <- read_tsv("source/intermediate/vic/ha/metadata_qc_location_clades_dedup_clades_glycans.tsv") |>
  mutate(
    date = if_else(
      str_detect(date, "2025-..-..") & !str_detect(date, "2025-\\d{2}-\\d{2}"),
      "2025-01-01", date
    ),
    date = as.Date(date)
  )

# Filter to relevant entries
nextclade_filt <- nextclade |> 
  filter(date >= as.Date("2022-01-01") & date <= as.Date("2025-12-31")) |> 
  filter(database_origin == "mostafa_lab") |> 
  mutate(ha_total_nglycans = as.numeric(ha_total_nglycans)) |> 
  filter(ha_total_nglycans >= 10) |> 
  drop_na(ha_total_nglycans, na_total_nglycans, subclade) |> 
  mutate(
    week = floor_date(date, "week"),
    ha_glycan_cat = case_when(
      ha_total_nglycans == 10 ~ "D/EET", 
      ha_total_nglycans == 11 ~ "NET",
      TRUE ~ as.character(ha_total_nglycans)  # fallback, just in case
    ),
    na_glycan_cat = as.character(na_total_nglycans)
  )

# Check if unwanted values remain
unique(nextclade_filt$ha_glycan_cat)

ha_glycan_freq <- nextclade_filt |> 
  count(week, ha_glycan_cat) |> 
  group_by(week) |> 
  mutate(freq = n / sum(n)) |> 
  ungroup()

# Ridge plot
p_ha_ridge <- ha_glycan_freq |> 
  ggplot(aes(x = week, y = n, fill = ha_glycan_cat)) +
  geom_stream(sorting = "inside_out", bw = 1.2, type = "ridge", color = "black") +
  scale_fill_manual(values = glycan_colors, name = "Site 197-199") +
  labs(x = "Sequencing Date", y = "Sequence count") +
  scale_x_date(date_labels = "%b \n %y", expand = expansion(mult = c(0.05, 0.1))) +
  scale_y_continuous(breaks = c(0, 50, 100, 150, 200, 250), limits = c(0,240)) +
  theme_classic() +
  theme(legend.position = "none", 
        axis.title = element_text(size = 18, face = "bold"),
        axis.text = element_text(size = 16),
        legend.text = element_text(size = 15),  # Increase legend text size
        legend.title = element_text(size = 15))  # Increase legend title size

# Proportional plot
p_ha_proportion <- ha_glycan_freq |>
  ggplot(aes(x = week, y = n, fill = ha_glycan_cat)) +
  geom_stream(sorting = "onset", bw = 1.2, type = "proportional", color = "black") +
  scale_fill_manual(values = glycan_colors, name = "Site 197-199") +
  labs(x = "Sequencing Date", y = "proportion") +
  scale_x_date(date_labels = "%b \n %y", expand = expansion(mult = c(0.05, 0.1))) +
  theme_classic() +  #
  theme(legend.position = "top",
        axis.title = element_text(size = 18, face = "bold"),
        axis.text = element_text(size = 16),
        legend.text = element_text(size = 15),  # Increase legend text size
        legend.title = element_text(size = 15))   # Increase legend title size

p_ha_ridge
p_ha_proportion

ggsave('figures/frequency_197_global_ridgeline.png', p_ha_ridge, height = 5, width = 5)
ggsave('figures/frequency_197_global_proportional.png', p_ha_proportion, height = 5, width = 5)
ggsave('figures/frequency_197_global_ridgeline-supp.png', p_ha_ridge, height = 4, width = 8)
ggsave('figures/frequency_197_global_proportional-supp.png', p_ha_proportion, height = 4, width = 8)


# try geom density
# Ridge plot


p_ha_ridge2 <- ha_glycan_freq |> 
  ggplot(aes(x = week, y = n, fill = ha_glycan_cat)) +
  #geom_stream(sorting = "inside_out", bw = 1.2, type = "ridge", color = "black") +
  geom_area() +
  scale_fill_manual(values = glycan_colors, name = "HA Glycans") +
  labs(x = "Date", y = "Sequence count", title = "HA Glycosylation Count Over Time") +
  theme_classic() +
  theme(axis.title = element_text(size = 14),
        axis.text = element_text(size = 12),
        legend.position = "top")



```

# Aggregated subclade table

```{r}
library(tidyverse)
library(gt)

nextclade_filt_table <- nextclade |> 
  filter(
    date >= as.Date("2024-10-01") & date <= as.Date("2025-12-31")
  )

# Filter and summarize
subclade_summary <- nextclade_filt_table |> 
  filter(!is.na(subclade)) |> 
  mutate(location_group = case_when(
    division == "Maryland" ~ "Maryland",
    region == "North America" ~ "North America",
    TRUE ~ NA_character_
  )) |> 
  filter(!is.na(location_group)) |> 
  count(location_group, subclade) |> 
  pivot_wider(names_from = location_group, values_from = n, values_fill = 0) |> 
  mutate(
    `Maryland Percent` = Maryland / sum(Maryland) * 100,
    `North America Percent` = `North America` / sum(`North America`) * 100
  ) |> 
  arrange(desc(`North America`)) |> 
  rename(
    `Subclade` = subclade,
    `Maryland Count` = Maryland,
    `North America Count` = `North America`
  )

# Format with gt
subclade_summary |> 
  gt() |> 
  fmt_number(columns = c(`North America Count`, `North America Percent`), decimals = 1) |> 
  cols_label(
    `Subclade` = "Subclade",
    `Maryland Count` = "Maryland Count",
    `Maryland Percent` = "Maryland Percent",
    `North America Count` = "North America Count",
    `North America Percent` = "North America Percent"
  ) |> 
  tab_header(
    title = "IBV Subclade Sequence Submissions\nin Maryland vs North America"
  )

table_subclade = subclade_summary |> 
  gt() |> 
  fmt_number(
    columns = c(`North America Count`, `Maryland Count`),
    decimals = 0
  ) |> 
  fmt_percent(
    columns = c(`North America Percent`, `Maryland Percent`),
    decimals = 1,
    scale_values = FALSE
  ) |> 
  cols_label(
    `Subclade` = "Subclade",
    `Maryland Count` = "Maryland Count",
    `Maryland Percent` = "Maryland Percent",
    `North America Count` = "North America Count",
    `North America Percent` = "North America Percent"
  ) |> 
  cols_align(
    align = "center",
    columns = everything()
  ) |> 
  tab_header(
    title = "IBV Subclade Sequence Submissions\nin Maryland vs North America"
  )

table_subclade |>
  gtsave("figures/table_subclade_frequency.docx")

table_subclade |>
  gtsave("figures/table_subclade_frequency.html")

```

# C.3 and C.3re distribution in JHH

```{r}
library(tidyverse)
library(ggstream)
library(plotly)
library(patchwork)

# import nextclade data 
nextclade <- read_tsv("source/intermediate/vic/ha/metadata_qc_location_clades_dedup_clades_glycans.tsv") |>
  mutate(date = if_else(str_detect(date, "2025-..-..") & !str_detect(date, "2025-\\d{2}-\\d{2}"),
                        "2025-01-01", date),
         date = as.Date(date))

#filter out vaccine strain
nextclade_filt = nextclade |> 
  filter(date >= as.Date("2022-01-01") & date <= as.Date("2025-12-31")) |>
  filter(database_origin == "mostafa_lab")

# Count occurrences of each subclade per date
frequency <-  nextclade_filt |> 
  count(date, subclade) |>
  ungroup()

# Generate sequence of dates representing each week
weekly_dates <- seq(min(frequency$date), max(frequency$date), by = "week")

monthly_dates <- seq(floor_date(min(frequency$date), "month"),
                     floor_date(max(frequency$date), "month"),
                     by = "month")

# Create the base plot for subclade frequency
p1 = frequency |> 
  ggplot(aes(date, n, fill=subclade), color = "black") + 
  scale_fill_manual(values = c(subclade_colors, "gray")) +
  theme_classic()

ridge = p1 +
  geom_stream(sorting = "inside_out", bw = 1.2, type = "ridge", color = "black",) +
  #geom_vline(xintercept = as.numeric(monthly_dates), linetype = "solid", color = "black", alpha = 0.1) +
  scale_x_date(date_labels = "%b \n %y", expand = expansion(mult = c(0.05, 0.1))) +  #
  theme(legend.position = "none",
        axis.title = element_text(size = 18, face = "bold"),
        axis.text = element_text(size = 16),
        legend.text = element_text(size = 15),  # Increase legend text size
        legend.title = element_text(size = 15)) +  # Increase legend title size
  labs(y="Sequence count", x="Sequencing Date")

proportional = p1 +
  geom_stream(sorting = "inside_out", bw = 1.2, type = "proportional", color = "black",) +
  #geom_vline(xintercept = as.numeric(monthly_dates), linetype = "solid", color = "black", alpha = 0.1) +
  scale_x_date(date_labels = "%b \n %y", expand = expansion(mult = c(0.05, 0.1))) +  #
  theme(legend.position = "top",
        axis.title = element_text(size = 18, face = "bold"),
        axis.text = element_text(size = 16),
        legend.text = element_text(size = 15),  # Increase legend text size
        legend.title = element_text(size = 15)) +  # Increase legend title size
  labs(y="Proportion", x="Sequencing Date")

# Create the base plot for glycosylation frequency

p2 = 

ggsave('figures/jhh_frequency_ridge.png', ridge, height = 4, width = 4)
ggsave('figures/jhh_supplemental_frequency_ridge.png', ridge, height = 4, width = 8)
ggsave('figures/jhh_supplemental_frequency_proportional.png', proportional, height = 4, width = 8)

ridge
proportional
```

## Country Level

```{r}
library(tidyverse)

nextstrain <- read_tsv("source/intermediate/vic/ha/metadata_qc_location_clades_dedup_clades_glycans.tsv")
```

# Multiple Allele Frequencies...

Using the non-downsampled nextclade-datasets, we need to extract specific residue coordinates , their alleles and populate a column for frequency calcluation and plotting.

our table is source/intermediate/vic/ha/metadata_qc_location_clades_dedup_clades_glycans.tsv and I have imported it with nextclade \<- read_tsv("source/intermediate/vic/ha/metadata_qc_location_clades_dedup_clades_glycans.tsv"). Here are the colnames

> colnames(nextstrain) \[1\] "sequence_ID" "sample_ID" "type"\
> \[4\] "subtype" "date" "passage_history"\
> \[7\] "study_id" "sequencing_run" "database_origin"\
> \[10\] "age" "age_unit" "sex"\
> \[13\] "seqName" "ha_glycosylation" "na_glycosylation"\
> \[16\] "genome_qc.overallScore" "genome_qc.overallStatus" "region"\
> \[19\] "country" "division" "area"\
> \[22\] "clade" "subclade" "ha_total_nglycans"\
> \[25\] "na_total_nglycans"

Step 1: Import amino acid substitutions columns:

ha_aaSubstitutions \<- read_tsv("source/intermediate/vic/ha/nextclade.tsv", only include the seqName column and aaSubstitutions column) na_aaSubstitutions \<- read_tsv("source/intermediate/vic/na/nextclade.tsv", only include the seqName column and aaSubstitutions column)

Step 2: Left join the substitution tables one after another onto the nexclade master table. The column to join on the nextstrain table is sequence_ID and the column on both of the subsstitution tables is seqName. Perform a left join onto the nextclade table to only retain rows present in that dataframe. Call the 2 column ha_aaSubstitutions, and ha_aaSubstitutions.

Step 3: Here's where things get tricky. Amino acid substitutions are encoded in both of the joined columns by the coding sequence followed by a colon, the reference allele, the relative residue coordinate poisiton, and the alternate allele. Different substitutions in the same row will be seperated by a comma (,). Additionally, different rows will have different amounts of alleles

An example of an allele NA:I120V means on the NA coding sequence, position 120 is now a V (valine) where it is a Isoleucine in the reference. Here is what several alleles look like in each row. Each row and will have multple substitusion and coding sequences so each of these components needs to be parsed individually later...

NA:P42Q,NA:L52S,NA:V71A,NA:I120V,NA:K220N,NA:S295R,NA:N340D,NA:K343E,NA:E358K,NA:G378E,NA:D384G,NA:A395I,NA:V401I,NB:F3L

I want to be able to specifically reference single or multiple coding sequence and residue position from these column and extract out the alternate alleles for each for frequecny quantifcation against the date column.

For example, I want to be able to specify NA positions 186 and 395 then get a frequency plot over time which shows abundance of the alternate alles for both catagorically over time. The cataogry would appear as the alt/alt alleles for each coding sequence. If I only reference 1 coding sequence, only show one. I want it to be dynamic to theoretically handle 1 or multiple coding sequences, positions and alleles while maintinating this information when plotting.

## Function to plot allele frequencies

```{r}
get_alt_freq_plot <- function(parsed_subs, target_sites, time_var = "date") {
  # target_sites: list like list(NA = c(186, 395), HA = c(197))
  
  # Convert to tibble for joining
  target_df <- enframe(target_sites, name = "gene", value = "position") %>%
    unnest(position)

  # Filter to target sites
  filtered <- parsed_subs %>%
    semi_join(target_df, by = c("gene", "position"))

  # Construct allele label like "NA:186V"
  filtered <- filtered %>%
    mutate(mutation = paste0(gene, ":", position, alt))

  # Count frequency by date
  freq_table <- filtered %>%
    group_by(across(all_of(time_var)), mutation) %>%
    summarise(count = n(), .groups = "drop")

  # Total sequences per timepoint (for normalization)
  total_by_date <- nextclade_joined %>%
    count(across(all_of(time_var))) %>%
    rename(total = n)

  # Join and calculate frequency
  freq_table <- freq_table %>%
    left_join(total_by_date, by = time_var) %>%
    mutate(freq = count / total)

  # Plot
  p <- ggplot(freq_table, aes_string(x = time_var, y = "freq", fill = "mutation")) +
    geom_col(position = "stack") +
    labs(y = "Frequency", x = time_var, fill = "Mutation") +
    theme_minimal()
  
  return(p)
}
```

## Import and clean data

```{r}
library(tidyverse)
library(lubridate)

# Step 1: Load main nextclade.tsv
nextclade <- read_tsv("source/intermediate/vic/ha/metadata_qc_location_clades_dedup_clades_glycans.tsv")

# Step 2: Load HA and NA substitutions
ha_aaSubstitutions <- read_tsv("source/intermediate/vic/ha/nextclade.tsv",
                               col_select = c(seqName, aaSubstitutions)) %>%
  rename(ha_aaSubstitutions = aaSubstitutions)

na_aaSubstitutions <- read_tsv("source/intermediate/vic/na/nextclade.tsv",
                               col_select = c(seqName, aaSubstitutions)) %>%
  rename(na_aaSubstitutions = aaSubstitutions)

# Step 3: Join both HA and NA substitutions onto metadata table
nextclade_joined <- nextclade %>%
  left_join(ha_aaSubstitutions, by = c("sample_ID" = "seqName")) %>%
  left_join(na_aaSubstitutions, by = c("sample_ID" = "seqName")) %>%
  filter(database_origin == "mostafa_lab")

# Step 4: Fix dates with 'XX' month or day — replace with April 1st
nextclade_joined <- nextclade_joined %>%
  mutate(date = str_replace_all(date, "-XX", "-04"),  # Replace unknown MM or DD with '04'
         date = suppressWarnings(ymd(date)))          # Parse cleaned date

# Step 5: Combine HA and NA substitutions into long format for parsing
parsed_subs <- nextclade_joined %>%
  select(sequence_ID, date, ha_aaSubstitutions, na_aaSubstitutions) %>%
  pivot_longer(cols = c(ha_aaSubstitutions, na_aaSubstitutions),
               names_to = "source", values_to = "aaSubstitutions") %>%
  filter(!is.na(aaSubstitutions)) %>%
  separate_rows(aaSubstitutions, sep = ",") %>%
  extract(aaSubstitutions, into = c("gene", "ref", "position", "alt"),
          regex = "([^:]+):([A-Z])([0-9]+)([A-Z])", convert = TRUE) %>%
  mutate(gene = case_when(
    is.na(gene) & source == "na_aaSubstitutions" ~ "NA",  # ← 🔧 fix NA → "NA"
    TRUE ~ gene
  ))

```

# New function for smoothed multi frequencies

```{r}
# WARNING - this exclusively summarizes the observed frequencies in the dataframe. This needs to be rewritten 

plot_mutation_combo_counts <- function(parsed_subs, target_sites, time_var = "date") {
  # Step 1: Build target position table
  target_df <- enframe(target_sites, name = "gene", value = "position") %>%
    unnest(position)
  
  # Step 2: Filter substitutions to only those positions
  filtered <- parsed_subs %>%
    semi_join(target_df, by = c("gene", "position"))
  
  # Step 3: For each sequence, collapse genotype string (e.g., K/I or K/I + S/R)
  combos <- filtered %>%
    mutate(genotype = paste0(ref, "/", alt)) %>%
    arrange(sequence_ID, gene, position) %>%
    group_by(sequence_ID, !!sym(time_var)) %>%
    summarise(genotype_combo = paste(genotype, collapse = " + "), .groups = "drop")

  # Step 4: Count each genotype combo by time
  counts <- combos %>%
    group_by(!!sym(time_var), genotype_combo) %>%
    summarise(count = n(), .groups = "drop")
  
  # Step 5: Plot
  ggplot(counts, aes_string(x = time_var, y = "count", fill = "genotype_combo")) +
    geom_area(position = "stack") +
    labs(
      y = "Number of sequences",
      x = time_var,
      fill = "Genotype",
      title = paste0("Genotype at ", 
                     paste(paste(names(target_sites), target_sites, sep = " site "), collapse = ", "))
    ) +
    theme_minimal()
}

```

## test new function plots

```{r}
# NA 186 and 295, raw counts over time
target_sites <- list("NA" = c(186, 395))
plot_mutation_combo_counts(parsed_subs, target_sites)

```

# Apply multi function with streagraph library

```{r}
library(tidyverse)
library(ggstream)

plot_mutation_combo_counts <- function(parsed_subs, target_sites, time_var = "date") {
  # Step 1: Build target position table
  target_df <- enframe(target_sites, name = "gene", value = "position") %>%
    unnest(position)
  
  # Step 2: Filter substitutions to only those positions
  filtered <- parsed_subs %>%
    semi_join(target_df, by = c("gene", "position"))
  
  # Step 3: For each sequence, collapse genotype string (e.g., K/I or K/I + S/R)
  combos <- filtered %>%
    mutate(genotype = paste0(ref, "/", alt)) %>%
    arrange(sequence_ID, gene, position) %>%
    group_by(sequence_ID, !!sym(time_var)) %>%
    summarise(genotype_combo = paste(genotype, collapse = " + "), .groups = "drop")

  # Step 4: Count each genotype combo by time
  counts <- combos %>%
    group_by(!!sym(time_var), genotype_combo) %>%
    summarise(count = n(), .groups = "drop")
  
  # Step 5: Plot
  ggplot(counts, aes_string(x = time_var, y = "count", fill = "genotype_combo")) +
    geom_stream(bw = 1.2, type = c("ridge"), sorting = c("inside_out")) +
    labs(
      y = "Number of sequences",
      x = time_var,
      fill = "Genotype",
      title = paste0("Genotype at ", 
                     paste(paste(names(target_sites), target_sites, sep = " site "), collapse = ", "))
    ) +
    theme_minimal()
}
```

## test new function plots

```{r}
# NA 186 and 295, raw counts over time
target_sites <- list("NA" = c(186, 395))
plot_mutation_combo_counts(parsed_subs, target_sites)
```

# FINAL FUNCTION

```{r}
library(tidyverse)
library(ggstream)

plot_mutation_combo_counts <- function(parsed_subs, target_sites, time_var = "date") {
  # Step 1: Build target position table
  target_df <- enframe(target_sites, name = "gene", value = "position") %>%
    unnest(position)
  
  # Step 2: Filter substitutions to only those positions
  filtered <- parsed_subs %>%
    semi_join(target_df, by = c("gene", "position"))
  
  # Step 3: For each sequence, collapse genotype string (e.g., K/I or K/I + S/R)
  combos <- filtered %>%
    mutate(genotype = paste0(ref, "/", alt)) %>%
    arrange(sequence_ID, gene, position) %>%
    group_by(sequence_ID, !!sym(time_var)) %>%
    summarise(genotype_combo = paste(genotype, collapse = " + "), .groups = "drop")

  # Step 4: Count each genotype combo by time
  counts <- combos %>%
    group_by(!!sym(time_var), genotype_combo) %>%
    summarise(count = n(), .groups = "drop")

  # Step 5: Build dynamic legend label
  legend_label <- target_df %>%
    arrange(gene, position) %>%
    group_by(gene) %>%
    summarise(sites = paste(position, collapse = ", "), .groups = "drop") %>%
    mutate(label = paste0(gene, " site ", sites)) %>%
    pull(label) %>%
    paste(collapse = " and ")

  legend_label <- paste("Genotype at \n", legend_label)

  # Step 6: Plot using ggstream
  ggplot(counts, aes_string(x = time_var, y = "count", fill = "genotype_combo")) +
    geom_stream(bw = 1.2, type = "ridge", sorting = "inside_out") +
    #ylim(0,200) +
    #geom_area() +
    labs(
      y = "Number of sequences",
      x = time_var,
      fill = legend_label,
    ) +
    theme_classic() +
    scale_x_date(date_labels = "%b \n %y", expand = expansion(mult = c(0.05, 0.1))) +
    theme(legend.position = "right",
          axis.title = element_text(size = 18, face = "bold"),
          axis.text = element_text(size = 16),
          legend.text = element_text(size = 15),  # Increase legend text size
          legend.title = element_text(size = 15)) +  # Increase legend title size
    labs(y="Sequence count", x="Sequencing Date")
}

```

## test labeling

```{r}

target_sites <- list("NA" = c(186,395))
plot_mutation_combo_counts(parsed_subs, target_sites)

target_sites <- list("HA1" = c(197,198,199))
plot_mutation_combo_counts(parsed_subs, target_sites)
```

Okay this script does exactly what its supposed to do. However, many sequences are being left out of the count IF they encode the reference allele and not the alternate allele...

# Missing data compatible function with ggstream

```{r}
plot_mutation_combo_counts <- function(parsed_subs, target_sites, time_var = "date") {
  library(tidyverse)
  library(ggstream)

  # Step 1: Build target position table
  target_df <- enframe(target_sites, name = "gene", value = "position") %>%
    unnest(position)

  # Step 2: Filter substitutions to only those positions
  filtered <- parsed_subs %>%
    semi_join(target_df, by = c("gene", "position"))

  # Step 3: Collapse genotype strings
  combos <- filtered %>%
    mutate(genotype = paste0(ref, "/", alt)) %>%
    arrange(sequence_ID, gene, position) %>%
    group_by(sequence_ID, !!sym(time_var)) %>%
    summarise(genotype_combo = paste(genotype, collapse = " + "), .groups = "drop")

  # Step 4: Drop rows with missing date
  combos <- combos %>%
    filter(!is.na(!!sym(time_var)))

  # Step 5: Count genotype combos per time
  counts <- combos %>%
    group_by(!!sym(time_var), genotype_combo) %>%
    summarise(count = n(), .groups = "drop")

  # Step 6: Fill in missing time/genotype combos with count = 0
  all_dates <- unique(combos[[time_var]])
  all_genotypes <- unique(counts$genotype_combo)
  full_grid <- expand_grid(!!sym(time_var) := all_dates, genotype_combo = all_genotypes)

  counts_filled <- full_grid %>%
    left_join(counts, by = c(time_var, "genotype_combo")) %>%
    mutate(count = replace_na(count, 0))

  # Step 7: Build legend label
  legend_label <- target_df %>%
    arrange(gene, position) %>%
    group_by(gene) %>%
    summarise(sites = paste(position, collapse = ", "), .groups = "drop") %>%
    mutate(label = paste0(gene, " site ", sites)) %>%
    pull(label) %>%
    paste(collapse = " and ")

  legend_label <- paste("Genotype at", legend_label)

  # Step 8: Plot
  #ggplot(counts_filled, aes_string(x = time_var, y = "count", fill = "genotype_combo")) +
  ggplot(counts_filled, aes_string(x = time_var, y = "count", fill = "genotype_combo", group = "genotype_combo")) +
    geom_stream(bw = 1.2, type = "ridge", sorting = "inside_out") +
    labs(
      y = "Number of sequences",
      x = time_var,
      fill = legend_label,
      title = legend_label
    ) +
    theme_minimal()
}

```

## test labeling

```{r}
# One gene, multiple sites
target_sites <- list("NA" = c(186, 395))
plot_mutation_combo_counts(parsed_subs, target_sites)

```

# TROUBLESHOOTING Diagnosing missing values


```{r}
library(tidyverse)

# --- Setup ---
target_df <- enframe(target_sites, name = "gene", value = "position") %>%
  unnest(position)

# Filter parsed_subs for sites of interest
filtered <- parsed_subs %>%
  semi_join(target_df, by = c("gene", "position"))

# Collapse genotypes for each sequence and date
combos <- filtered %>%
  mutate(genotype = paste0(ref, "/", alt)) %>%
  arrange(sequence_ID, gene, position) %>%
  group_by(sequence_ID, date) %>%
  summarise(genotype_combo = paste(genotype, collapse = " + "), .groups = "drop")

# Count occurrences of each genotype per date
counts <- combos %>%
  group_by(date, genotype_combo) %>%
  summarise(count = n(), .groups = "drop")

counts %>%
  group_by(genotype_combo) %>%
  summarise(n_dates = n_distinct(date)) %>%
  arrange(n_dates) %>%
  print(n = 100)

counts %>%
  group_by(date) %>%
  summarise(n_genos = n_distinct(genotype_combo)) %>%
  arrange(n_genos)

all_dates <- sort(unique(counts$date))
all_genotypes <- sort(unique(counts$genotype_combo))

expected <- expand_grid(date = all_dates, genotype_combo = all_genotypes)

missing_combos <- anti_join(expected, counts, by = c("date", "genotype_combo"))

if (nrow(missing_combos) == 0) {
  message("✅ All combinations are present!")
} else {
  message("❌ Missing combinations found: ", nrow(missing_combos))
  print(missing_combos, n = 20)
}


```

```{r}
counts %>%
  mutate(present = ifelse(count > 0, 1, 0)) %>%
  ggplot(aes(x = date, y = genotype_combo, fill = factor(present))) +
  geom_tile(color = "white") +
  scale_fill_manual(values = c("0" = "grey90", "1" = "steelblue")) +
  labs(
    title = "Presence matrix: date × genotype_combo",
    x = "Date", y = "Genotype",
    fill = "Present"
  ) +
  theme_minimal(base_size = 12)

```

```{r}
counts_filled <- expand_grid(date = all_dates, genotype_combo = all_genotypes) %>%
  left_join(counts, by = c("date", "genotype_combo")) %>%
  mutate(count = replace_na(count, 0)) %>%
  arrange(date, genotype_combo)

ggplot(counts_filled, aes(x = date, y = count, fill = genotype_combo, group = genotype_combo)) +
  geom_stream(bw = 1.2, type = "ridge", sorting = "inside_out") +
  theme_minimal()


```

# Test NEW Multi Allele Function

```{r}
library(tidyverse)
library(Biostrings)
library(ggplot2)
library(ggnewscale)

# Helper function: parse mixed position input
parse_position_input <- function(pos_input) {
  if (is.null(pos_input)) return(NULL)
  
  ranges <- str_split(pos_input, ",\\s*")[[1]]
  pos_list <- c()
  for (r in ranges) {
    if (str_detect(r, "-")) {
      bounds <- as.integer(str_split(r, "-")[[1]])
      pos_list <- c(pos_list, seq(bounds[1], bounds[2]))
    } else {
      pos_list <- c(pos_list, as.integer(r))
    }
  }
  sort(unique(pos_list))
}

# Helper: Convert alignment to tidy tibble
alignment_to_tidy <- function(aln_file) {
  aln <- readDNAStringSet(aln_file)
  seq_names <- names(aln)
  seqs <- as.character(aln)
  pos_df <- do.call(rbind, strsplit(seqs, split = ""))
  pos_df <- as_tibble(pos_df)
  pos_df$seqname <- seq_names
  pivot_longer(pos_df, -seqname, names_to = "position", values_to = "base") |>
    mutate(position = as.integer(position))
}

# Helper: Compute allele frequencies
compute_freqs <- function(aln_tidy, metadata_df, positions = NULL, date_range = NULL) {
  # Join with metadata
  df <- aln_tidy |>
    inner_join(metadata_df, by = c("SeqName" = "sequence_ID")) |>
    mutate(date = lubridate::ymd(date)) 
  
  if (!is.null(date_range)) {
    df <- df |> filter(date >= date_range[1] & date <= date_range[2])
  }
  if (!is.null(positions)) {
    df <- df |> filter(position %in% positions)
  }

  df |>
    count(date, position, base, name = "obs_count") |>
    group_by(date, position) |>
    mutate(freq = obs_count / sum(obs_count)) |>
    ungroup()
}

# Final wrapper function
plot_allele_frequencies <- function(
  aln_file,
  metadata_file,
  pos_input = NULL,            # e.g. "155,158,161-163"
  date_range = NULL,           # e.g. c("2023-10-01", "2024-03-30")
  color_scheme = NULL          # named vector like c("A" = "red", "G" = "blue", ...)
) {
  metadata_df <- readr::read_delim(metadata_file, show_col_types = FALSE) |>
    mutate(date = lubridate::ymd(date))

  positions <- parse_position_input(pos_input)
  aln_tidy <- alignment_to_tidy(aln_file)
  freq_df <- compute_freqs(aln_tidy, metadata_df, positions, date_range)

  base_colors <- if (!is.null(color_scheme)) color_scheme else RColorBrewer::brewer.pal(8, "Dark2")

  ggplot(freq_df, aes(x = date, y = freq, fill = base)) +
    geom_area(position = "stack", color = "white", linewidth = 0.25) +
    scale_fill_manual(values = base_colors) +
    facet_wrap(~ position, scales = "free_y", ncol = 1) +
    scale_x_date(date_breaks = "1 month", date_labels = "%b\n%y") +
    labs(x = "Date", y = "Allele Frequency", fill = "Allele") +
    theme_classic(base_size = 14) +
    theme(
      strip.text = element_text(size = 13, face = "bold"),
      axis.text.x = element_text(angle = 0, hjust = 0.5),
      panel.spacing.y = unit(1.5, "lines")
    )
}
```

## Test function 

HA 197-199
```{r}
plot_allele_frequencies(
  aln_file = "results/ha_nextclade_complete/nextclade.cds_translation.HA1_jhhonly.fasta",
  metadata_file = "results/ha_nextclade_complete/filtered.tsv",
  date_range = c("2023-10-01", "2025-05-01"),
  pos_input = "197"
)
```

## Try again
```{r}
library(Biostrings)
library(tidyverse)
library(lubridate)

plot_allele_frequencies <- function(
  fasta_file,
  metadata_file,
  positions = NULL,
  ranges = NULL,
  date_range = NULL,
  color_map = NULL
) {
  # --- Step 1: Parse alignment ---
  alignment <- readAAStringSet(fasta_file)
  seq_names <- names(alignment)
  seq_matrix <- as.matrix(alignment)
  colnames(seq_matrix) <- paste0("pos_", seq(1, ncol(seq_matrix)))
  df <- as.data.frame(seq_matrix)
  df <- df %>%
    mutate(sequence_ID = seq_names) %>%
    relocate(sequence_ID)
  
  # --- Step 2: Select positions and/or ranges ---
  selected_positions <- c()
  if (!is.null(positions)) {
    selected_positions <- c(selected_positions, positions)
  }
  if (!is.null(ranges)) {
    range_positions <- unlist(lapply(ranges, function(r) seq(r[1], r[2])))
    selected_positions <- c(selected_positions, range_positions)
  }
  selected_positions <- sort(unique(selected_positions))
  selected_cols <- paste0("pos_", selected_positions)
  
  if (!all(selected_cols %in% colnames(df))) {
    stop("One or more selected positions are not in the alignment.")
  }
  
  long_df <- df %>%
    select(sequence_ID, all_of(selected_cols)) %>%
    pivot_longer(-sequence_ID, names_to = "position", values_to = "allele") %>%
    mutate(position = as.integer(gsub("pos_", "", position)))
  
  # --- Step 3: Merge positions if more than one ---
  if (length(selected_positions) > 1) {
    combo_df <- long_df %>%
      arrange(sequence_ID, position) %>%
      group_by(sequence_ID) %>%
      summarize(combined_allele = paste0(allele, collapse = ""), .groups = "drop")
  } else {
    combo_df <- long_df %>%
      rename(combined_allele = allele) %>%
      select(sequence_ID, combined_allele)
  }
  
  # --- Step 4: Join metadata ---
  metadata <- read_tsv(metadata_file, col_types = cols()) %>%
      select(sequence_ID, date) %>%
      mutate(
        date = if_else(
          str_detect(date, "^2025-..-..$") & !str_detect(date, "^2025-\\d{2}-\\d{2}$"),
          "2025-01-01",
          date
        ),
        date = as.Date(date)
      )
  
  if (!"sequence_ID" %in% colnames(metadata)) {
    stop("The metadata file must contain a 'sequence_ID' column.")
  }
  if (!"date" %in% colnames(metadata)) {
    stop("The metadata file must contain a 'date' column.")
  }

  # Make sure date is parsed correctly
  metadata <- metadata %>%
    mutate(date = ymd(date))

  # Apply date filter if specified
  if (!is.null(date_range)) {
    metadata <- metadata %>%
      filter(date >= ymd(date_range[1]) & date <= ymd(date_range[2]))
  }
  
  combined_df <- inner_join(combo_df, metadata, by = "sequence_ID")
  
  # --- Step 5: Calculate frequency counts ---
  freq_df <- combined_df %>%
    group_by(date, combined_allele) %>%
    summarise(n = n(), .groups = "drop")
  
  # Ensure dates are sorted and complete
  freq_df <- freq_df %>%
    arrange(date)

  # --- Step 6: Plot ---
  p <- ggplot(freq_df, aes(x = date, y = n, fill = combined_allele)) +
    geom_stream(bw = 1.2, type = "ridge", sorting = "inside_out") +
    scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
    scale_x_date(date_labels = "%b\n%y", expand = expansion(mult = c(0.01, 0.05))) +
    labs(x = "Collection Date", y = "Observation Count", fill = "Allele") +
    theme_classic(base_size = 14) +
    theme(legend.position = "right", axis.title = element_text(face = "bold")
    )

  if (!is.null(color_map)) {
    p <- p + scale_fill_manual(values = color_map)
  }

  return(p)
}
```


```{r}
plot_allele_frequencies(
  fasta_file = "results/ha_nextclade_complete/nextclade.cds_translation.HA1_jhhonly.fasta",
  metadata_file = "results/ha_nextclade_complete/filtered.tsv",
  positions = c(197,198,199),
  date_range = c("2024-10-01", "2025-05-01"),
  #color_map = c("SNA" = "#1b9e77", "TNA" = "#d95f02", "TNS" = "#7570b3")
)
```