---
title: "Observational Case Frequencies"
author: Elgin Akin
date: 2025-06-15
format: html
---


- use the `source/intermediate/vic/ha/metadata_qc_location_clades_dedup_clades_glycans.tsv` as the input data frame
- de-duplicated by strain name
- location column expanded to region / country / division / location
    - will need to look at the number of missing rows within each category...


# Figures 

## Total Frequency Summarized by Month
1. Global Frequency by Subclade
2. North American Frequency by subclade
3. Baltimore/JHH Frequency by Subclade
4. D197E Glycosylation by Month 

## Maps Global Distribution 

1. C.3 distribution globally
2. C.5.1 distribution globally
3. C.5.6 distribution globally
4. C.5.7 distribution globally


# Config

```{r}
subclade_colors <- c("C.5.1" = "#d95f02", 
                     "C.5.5" = "#66c2a5", 
                     "C.5.6" = "#051334",
                     "C.5.7" = "#666666",
                     "C.5" = "#e6ab02", 
                     "C.3" = "#00A7E1")

subclade_colors <- c("C.5.7" = "#666666",
                      "C.5.6" = "#051334",
                      "C.5.5" = "#66c2a5",
                      "C.5.4" = "#575482",
                      "C.5.3" = "#7570b3",
                      "C.5.2" = "#9893E1",
                      "C.5.1" = "#d95f02",
                      "C.5" = "#623C74", 
                      "C.4" = "#E4B19B",
                      "C.3" = "#00A7E1",
                      "C.2" = "#F2CACA",
                      "C.1" = "#F8CFCF",
                      "C" = "#FFE2D8",
                      "A" = "#F9DC5C",
                      "A.3.2" = "#FAE588",
                      "A.3.1.1" = "#FCEFB4",
                      "A.3" = "#FDF4CB")

#Other pallettes: 
 # - https://loading.io/color/feature/DragonBall-%E3%83%89%E3%83%A9%E3%82%B4%E3%83%B3%E3%83%9C%E3%83%BC%E3%83%AB

#Dark2
#1b9e77
#d95f02
#7570b3
#e7298a
#66a61e
#e6ab02
#a6761d
#666666

```


## Case Counts

```{r}
# Load necessary libraries
library(tidyverse)
library(ggplot2)
library(tidyr)
library(dplyr)
library(lubridate)

flu_data <- read_tsv("data/case_counts_2425.txt")

# Convert date column to Date type
flu_data$date_range <- as.Date(flu_data$date_range, format = "%Y-%m-%d")


# Reshape data to long format
flu_long <- flu_data %>%
  pivot_longer(cols = c(IAV_clinic, IBV_clinic), names_to = "Virus", values_to = "Case Count") %>%
  mutate(Virus = recode(Virus,
                        "IAV_clinic" = "Influenza A",
                        "IBV_clinic" = "Influenza B"))

                
# Filter to only show dates between October 2023 and April 2024
filtered_flu <- flu_long %>% filter(date_range >= as.Date("2024-10-02") & date_range <= as.Date("2025-12-31"))

# fill 2019-2024 df 
# summarized
filtered_fliu_monthly_dates <- seq(floor_date(min(filtered_flu$date_range), "month"),
                     floor_date(max(filtered_flu$date_range), "month"),
                     by = "month")

case_count = filtered_flu |>
  ggplot(aes(date_range, `Case Count`, color = Virus)) + 
  geom_smooth(se = FALSE, linewidth = 1.2) + 
  scale_x_date(date_labels = "%b\n%y",date_breaks = "1 month", expand = expansion(mult = c(0.05, 0.1))) +
  scale_color_manual(values = c("Influenza A" = "black", "Influenza B" = "red")) +
  labs(y = "Case Count", x = "Case Date", color = "Virus") +
  theme_classic(base_size = 16, base_family = "Helvetica") +
  theme(
    legend.position = "top",
    axis.title = element_text(size = 18, face = "bold"),
    axis.text = element_text(size = 16),
    legend.text = element_text(size = 15),
    legend.title = element_text(size = 15)
  )

case_count

ggsave('figures/case_count.png', case_count, height = 5, width = 5)

## save 




```


## Table of clinical cases by Month 
```{r}
library(tidyverse)
library(gt)

# Raw data
clinic_counts <- tibble(
  date_range = as.Date(c(
    "2024-10-01", "2024-11-01", "2024-12-01",
    "2025-01-01", "2025-02-01", "2025-03-01", "2025-04-01"
  )),
  IAV_clinic = c(2, 57, 510, 1988, 2229, 526, 53),
  IBV_clinic = c(0, 4, 12, 41, 65, 372, 400)
) |>
  mutate(
    Total = IAV_clinic + IBV_clinic,
    `IAV Percent` = IAV_clinic / Total * 100,
    `IBV Percent` = IBV_clinic / Total * 100
  ) |>
  rename(
    `Date` = date_range,
    `IAV Count` = IAV_clinic,
    `IBV Count` = IBV_clinic
  )

# Add grand total row
clinic_counts_totals <- clinic_counts |>
  summarise(
    Date = as.Date(NA),
    `IAV Count` = sum(`IAV Count`),
    `IBV Count` = sum(`IBV Count`),
    Total = sum(Total),
    `IAV Percent` = `IAV Count` / Total * 100,
    `IBV Percent` = `IBV Count` / Total * 100
  ) |>
  mutate(Date = as.Date("9999-12-31"))  # use dummy future date for sorting

# Bind totals row
final_table <- bind_rows(clinic_counts, clinic_counts_totals) |>
  arrange(Date) |>
  mutate(Date = if_else(Date == as.Date("9999-12-31"), as.Date(NA), Date))

# Format with gt
table_clinical_frequency = final_table |>
                                gt() |>
                                fmt_number(columns = c(`IAV Percent`, `IBV Percent`), decimals = 1) |>
                                #fmt_date(columns = "Date", date_style = "long") |>
                                fmt_date(columns = "Date", date_style = "yMMM") |>
                                cols_label(
                                  Date = "Date",
                                  `IAV Count` = "IAV Count",
                                  `IBV Count` = "IBV Count",
                                  Total = "Total Cases",
                                  `IAV Percent` = "IAV %",
                                  `IBV Percent` = "IBV %"
                                ) |>
                                tab_header(
                                  title = "Clinic-Detected Influenza Cases by Month",
                                  subtitle = "With Total IAV and IBV Counts"
                                ) |>
                                tab_style(
                                  style = cell_text(weight = "bold"),
                                  locations = cells_body(
                                    rows = is.na(Date)
                                  )
                                ) |>
                                text_transform(
                                  locations = cells_body(columns = "Date", rows = is.na(Date)),
                                  fn = function(x) "TOTAL"
                                )

table_clinical_frequency |> 
  gtsave("figures/table_clinic_frequency.docx")

table_clinical_frequency |> 
  gtsave("figures/table_clinic_frequency.pdf")

```

## Total Subclade Frequency Summarized by Month in JHH

```{r}
library(tidyverse)
library(ggstream)
library(plotly)
library(patchwork)

# import nextclade data 
nextclade <- read_tsv("source/intermediate/vic/ha/metadata_qc_location_clades_dedup_clades_glycans.tsv") |>
  mutate(date = if_else(str_detect(date, "2025-..-..") & !str_detect(date, "2025-\\d{2}-\\d{2}"),
                        "2025-01-01", date),
         date = as.Date(date))

#filter out vaccine strain
nextclade_filt = nextclade |> 
  filter(date >= as.Date("2022-01-01") & date <= as.Date("2025-12-31")) |>
  filter(database_origin == "mostafa_lab")

# Count occurrences of each subclade per date
frequency <-  nextclade_filt |> 
  count(date, subclade) |>
  ungroup()

# Generate sequence of dates representing each week
weekly_dates <- seq(min(frequency$date), max(frequency$date), by = "week")

monthly_dates <- seq(floor_date(min(frequency$date), "month"),
                     floor_date(max(frequency$date), "month"),
                     by = "month")

# Create the base plot for subclade frequency
p1 = frequency |> 
  ggplot(aes(date, n, fill=subclade), color = "black") + 
  scale_fill_manual(values = c(subclade_colors, "gray")) +
  theme_classic()

ridge = p1 +
  geom_stream(sorting = "inside_out", bw = 1.2, type = "ridge", color = "black",) +
  #geom_vline(xintercept = as.numeric(monthly_dates), linetype = "solid", color = "black", alpha = 0.1) +
  scale_x_date(date_labels = "%b \n %y", expand = expansion(mult = c(0.05, 0.1))) +  #
  theme(legend.position = "top",
        axis.title = element_text(size = 18, face = "bold"),
        axis.text = element_text(size = 16),
        legend.text = element_text(size = 15),  # Increase legend text size
        legend.title = element_text(size = 15)) +  # Increase legend title size
  labs(y="Sequence count", x="Sequencing Date")

proportional = p1 +
  geom_stream(sorting = "inside_out", bw = 1.2, type = "proportional", color = "black",) +
  #geom_vline(xintercept = as.numeric(monthly_dates), linetype = "solid", color = "black", alpha = 0.1) +
  scale_x_date(date_labels = "%b \n %y", expand = expansion(mult = c(0.05, 0.1))) +  #
  theme(legend.position = "top",
        axis.title = element_text(size = 18, face = "bold"),
        axis.text = element_text(size = 16),
        legend.text = element_text(size = 15),  # Increase legend text size
        legend.title = element_text(size = 15)) +  # Increase legend title size
  labs(y="Proportion", x="Sequencing Date")

# Create the base plot for glycosylation frequency

p2 = 

ggsave('figures/jhh_frequency_ridge.png', ridge, height = 5, width = 5)
ggsave('figures/jhh_supplemental_frequency_ridge.png', ridge, height = 4, width = 8)
ggsave('figures/jhh_supplemental_frequency_proportional.png', proportional, height = 4, width = 8)

ridge
proportional
```

### filter only to north america

```{r}
library(tidyverse)
library(ggstream)
library(plotly)
library(patchwork)


# import nextclade data 
nextclade <- read_tsv("source/intermediate/vic/ha/metadata_qc_location_clades_dedup_clades_glycans.tsv") |>
  mutate(date = if_else(str_detect(date, "2025-..-..") & !str_detect(date, "2025-\\d{2}-\\d{2}"),
                        "2025-01-01", date),
         date = as.Date(date))

#filter out vaccine strain
nextclade_filt <- nextclade |> 
  filter(
    date >= as.Date("2023-10-01") & date <= as.Date("2025-12-31"),
    database_origin != "mostafa_lab",
    country == "United States"
  )
  
# Count occurrences of each subclade per date
frequency <-  nextclade_filt |> 
  count(date, subclade) |>
  ungroup()

frequency <- nextclade_filt |>
  filter(!is.na(subclade), !is.na(date)) |>
  mutate(month = floor_date(date, unit = "month")) |>
  count(month, subclade) |>
  rename(date = month) |>
  ungroup()

# Generate sequence of dates representing each week
weekly_dates <- seq(min(frequency$date), max(frequency$date), by = "week")

monthly_dates <- seq(floor_date(min(frequency$date), "month"),
                     floor_date(max(frequency$date), "month"),
                     by = "month")

# Create the base plot
p1 = frequency |> 
  ggplot(aes(date, n, fill=subclade), color = "black") + 
  scale_fill_manual(values = c(subclade_colors, "gray")) +
  theme_classic()

ridge = p1 +
  geom_stream(sorting = "inside_out", bw = 1.2, type = "ridge", color = "black",) +
  #geom_vline(xintercept = as.numeric(monthly_dates), linetype = "solid", color = "black", alpha = 0.1) +
  scale_x_date(date_labels = "%b \n %y", expand = expansion(mult = c(0.05, 0.1))) +  #
  theme(legend.position = "top",
        axis.title = element_text(size = 18, face = "bold"),
        axis.text = element_text(size = 16),
        legend.text = element_text(size = 15),  # Increase legend text size
        legend.title = element_text(size = 15)) +  # Increase legend title size
  labs(y="Sequence count", x="Sequencing Date")

proportional = p1 +
  geom_stream(sorting = "inside_out", bw = 1.2, type = "proportional", color = "black",) +
  #geom_vline(xintercept = as.numeric(monthly_dates), linetype = "solid", color = "black", alpha = 0.1) +
  scale_x_date(date_labels = "%b \n %y", expand = expansion(mult = c(0.05, 0.1))) +  #
  theme(legend.position = "top",
        axis.title = element_text(size = 18, face = "bold"),
        axis.text = element_text(size = 16),
        legend.text = element_text(size = 15),  # Increase legend text size
        legend.title = element_text(size = 15)) +  # Increase legend title size
  labs(y="Proportion", x="Sequencing Date")

ggsave('figures/NHfrequency_ridgeline.png', ridge, height = 4, width = 8)
ggsave('figures/NHfrequency_proportional.png', proportional, height = 4, width = 8)

ridge
proportional
```


## Global 
```{r}
library(tidyverse)
library(ggstream)
library(plotly)
library(patchwork)

# import nextclade data 
nextclade <- read_tsv("source/intermediate/vic/ha/metadata_qc_location_clades_dedup_clades_glycans.tsv") |>
  mutate(date = if_else(str_detect(date, "2025-..-..") & !str_detect(date, "2025-\\d{2}-\\d{2}"),
                        "2025-01-01", date),
         date = as.Date(date))

#filter out vaccine strain
nextclade_filt <- nextclade |> 
  filter(
    date >= as.Date("2023-10-01") & date <= as.Date("2025-12-31"),
    database_origin != "mostafa_lab"
  )
  
# Count occurrences of each subclade per date
frequency <-  nextclade_filt |> 
  count(date, subclade) |>
  ungroup()

frequency <- nextclade_filt |>
  filter(!is.na(subclade), !is.na(date)) |>
  mutate(month = floor_date(date, unit = "month")) |>
  count(month, subclade) |>
  rename(date = month) |>
  ungroup()

# Generate sequence of dates representing each week
weekly_dates <- seq(min(frequency$date), max(frequency$date), by = "week")

monthly_dates <- seq(floor_date(min(frequency$date), "month"),
                     floor_date(max(frequency$date), "month"),
                     by = "month")

# Create the base plot
p1 = frequency |> 
  ggplot(aes(date, n, fill=subclade), color = "black") + 
  scale_fill_manual(values = c(subclade_colors, "gray")) +
  theme_classic()

ridge = p1 +
  geom_stream(sorting = "inside_out", bw = 1.2, type = "ridge", color = "black") +
  #geom_vline(xintercept = as.numeric(monthly_dates), linetype = "solid", color = "black", alpha = 0.1) +
  scale_x_date(date_labels = "%b \n %y", expand = expansion(mult = c(0.05, 0.1))) +  #
  theme(legend.position = "top",
        axis.title = element_text(size = 18, face = "bold"),
        axis.text = element_text(size = 16),
        legend.text = element_text(size = 15),  # Increase legend text size
        legend.title = element_text(size = 15)) +  # Increase legend title size
  labs(y="Sequence count", x="Sequencing Date")

proportional = p1 +
  geom_stream(sorting = "inside_out", bw = 1.2, type = "proportional", color = "black") +
  #geom_vline(xintercept = as.numeric(monthly_dates), linetype = "solid", color = "black", alpha = 0.1) +
  scale_x_date(date_labels = "%b \n %y", expand = expansion(mult = c(0.05, 0.1))) +  #
  theme(legend.position = "top",
        axis.title = element_text(size = 18, face = "bold"),
        axis.text = element_text(size = 16),
        legend.text = element_text(size = 15),  # Increase legend text size
        legend.title = element_text(size = 15)) +  # Increase legend title size
  labs(y="Proportion", x="Sequencing Date")

ggsave('figures/frequency_global_ridgeline.png', ridge, height = 4, width = 8)
ggsave('figures/frequency_global_proportional.png', proportional, height = 4, width = 8)

ridge
proportional
```



# Quantify N-linked glycosylation observation frequency

Use [quantify_hana_glycans.py](scripts/quantify_hana_glycans.py)

```{shell}
python scripts/quantify_hana_glycans.py \
  -i source/intermediate/vic/ha/metadata_qc_location_clades_dedup_clades_glycans.tsv \
  -o source/intermediate/vic/ha/metadata_qc_location_clades_dedup_clades_glycans.tsv \
  -ha ha_glycosylation \
  -na na_glycosylation
```

```{r}
library(tidyverse)
library(lubridate)


glycan_colors <- c(
    "EET" = "#72874EFF", 
    "NET" = "#FF746C"
)

library(tidyverse)
library(lubridate)

# Load data
nextclade <- read_tsv("source/intermediate/vic/ha/metadata_qc_location_clades_dedup_clades_glycans.tsv") |>
  mutate(
    date = if_else(
      str_detect(date, "2025-..-..") & !str_detect(date, "2025-\\d{2}-\\d{2}"),
      "2025-01-01", date
    ),
    date = as.Date(date)
  )

# Filter to relevant entries
nextclade_filt <- nextclade |> 
  filter(date >= as.Date("2022-01-01") & date <= as.Date("2025-12-31")) |> 
  filter(database_origin == "mostafa_lab") |> 
  mutate(ha_total_nglycans = as.numeric(ha_total_nglycans)) |> 
  filter(ha_total_nglycans >= 10) |> 
  drop_na(ha_total_nglycans, na_total_nglycans, subclade) |> 
  mutate(
    week = floor_date(date, "week"),
    ha_glycan_cat = case_when(
      ha_total_nglycans == 10 ~ "EET", 
      ha_total_nglycans == 11 ~ "NET",
      TRUE ~ as.character(ha_total_nglycans)  # fallback, just in case
    ),
    na_glycan_cat = as.character(na_total_nglycans)
  )

# Check if unwanted values remain
unique(nextclade_filt$ha_glycan_cat)

ha_glycan_freq <- nextclade_filt |> 
  count(week, ha_glycan_cat) |> 
  group_by(week) |> 
  mutate(freq = n / sum(n)) |> 
  ungroup()

# Ridge plot
p_ha_ridge <- ha_glycan_freq |> 
  ggplot(aes(x = week, y = n, fill = ha_glycan_cat)) +
  geom_stream(sorting = "inside_out", bw = 1.2, type = "ridge", color = "black") +
  scale_fill_manual(values = glycan_colors, name = "Site 197-199") +
  labs(x = "Sequencing Date", y = "Sequence count") +
  scale_x_date(date_labels = "%b \n %y", expand = expansion(mult = c(0.05, 0.1))) +  #
  theme_classic() +
  theme(legend.position = "top",
        axis.title = element_text(size = 18, face = "bold"),
        axis.text = element_text(size = 16),
        legend.text = element_text(size = 15),  # Increase legend text size
        legend.title = element_text(size = 15))  # Increase legend title size

# Proportional plot
p_ha_proportion <- ha_glycan_freq |>
  ggplot(aes(x = week, y = n, fill = ha_glycan_cat)) +
  geom_stream(sorting = "onset", bw = 1.2, type = "proportional", color = "black") +
  scale_fill_manual(values = glycan_colors, name = "Site 197-199") +
  labs(x = "Sequencing Date", y = "proportion") +
  scale_x_date(date_labels = "%b \n %y", expand = expansion(mult = c(0.05, 0.1))) +
  theme_classic() +  #
  theme(legend.position = "top",
        axis.title = element_text(size = 18, face = "bold"),
        axis.text = element_text(size = 16),
        legend.text = element_text(size = 15),  # Increase legend text size
        legend.title = element_text(size = 15))   # Increase legend title size

p_ha_ridge
p_ha_proportion

ggsave('figures/frequency_197_global_ridgeline.png', p_ha_ridge, height = 5, width = 5)
ggsave('figures/frequency_197_global_proportional.png', p_ha_proportion, height = 5, width = 5)
ggsave('figures/frequency_197_global_ridgeline-supp.png', p_ha_ridge, height = 4, width = 8)
ggsave('figures/frequency_197_global_proportional-supp.png', p_ha_proportion, height = 4, width = 8)


# try geom density
# Ridge plot


p_ha_ridge2 <- ha_glycan_freq |> 
  ggplot(aes(x = week, y = n, fill = ha_glycan_cat)) +
  #geom_stream(sorting = "inside_out", bw = 1.2, type = "ridge", color = "black") +
  geom_area() +
  scale_fill_manual(values = glycan_colors, name = "HA Glycans") +
  labs(x = "Date", y = "Sequence count", title = "HA Glycosylation Count Over Time") +
  theme_classic() +
  theme(axis.title = element_text(size = 14),
        axis.text = element_text(size = 12),
        legend.position = "top")



```



# Aggregated subclade table
```{r}
library(tidyverse)
library(gt)

nextclade_filt_table <- nextclade |> 
  filter(
    date >= as.Date("2024-10-01") & date <= as.Date("2025-12-31")
  )

# Filter and summarize
subclade_summary <- nextclade_filt_table |> 
  filter(!is.na(subclade)) |> 
  mutate(location_group = case_when(
    division == "Maryland" ~ "Maryland",
    region == "North America" ~ "North America",
    TRUE ~ NA_character_
  )) |> 
  filter(!is.na(location_group)) |> 
  count(location_group, subclade) |> 
  pivot_wider(names_from = location_group, values_from = n, values_fill = 0) |> 
  mutate(
    `Maryland Percent` = Maryland / sum(Maryland) * 100,
    `North America Percent` = `North America` / sum(`North America`) * 100
  ) |> 
  arrange(desc(`North America`)) |> 
  rename(
    `Subclade` = subclade,
    `Maryland Count` = Maryland,
    `North America Count` = `North America`
  )

# Format with gt
subclade_summary |> 
  gt() |> 
  fmt_number(columns = c(`North America Count`, `North America Percent`), decimals = 1) |> 
  cols_label(
    `Subclade` = "Subclade",
    `Maryland Count` = "Maryland Count",
    `Maryland Percent` = "Maryland Percent",
    `North America Count` = "North America Count",
    `North America Percent` = "North America Percent"
  ) |> 
  tab_header(
    title = "IBV Subclade Sequence Submissions\nin Maryland vs North America"
  )

table_subclade = subclade_summary |> 
  gt() |> 
  fmt_number(
    columns = c(`North America Count`, `Maryland Count`),
    decimals = 0
  ) |> 
  fmt_percent(
    columns = c(`North America Percent`, `Maryland Percent`),
    decimals = 1,
    scale_values = FALSE
  ) |> 
  cols_label(
    `Subclade` = "Subclade",
    `Maryland Count` = "Maryland Count",
    `Maryland Percent` = "Maryland Percent",
    `North America Count` = "North America Count",
    `North America Percent` = "North America Percent"
  ) |> 
  cols_align(
    align = "center",
    columns = everything()
  ) |> 
  tab_header(
    title = "IBV Subclade Sequence Submissions\nin Maryland vs North America"
  )

table_subclade |>
  gtsave("figures/table_subclade_frequency.docx")

table_subclade |>
  gtsave("figures/table_subclade_frequency.html")

```


# C.3 and C.3re distribution 

We can use the non-down selected C.3re nextclade data to get true subclade frequencies 


## Country Level

```{r}
library(tidyverse)

nextstrain <- read.tsv(source/intermediate/vic/ha/metadata_qc_location_clades_dedup_clades_glycans.tsv)
```