---
title: "2024-25 JHH Influenza Serology"
date: 2025-07-14
author: Elgin Akin
format: html
---

# TODO

# Notebook Description

This notebook houses NT50 data for NH2024-25 Vaccine and Circulating Influenza viruses Isolated from the JHH network using sera collected from the 24-25 CEIRR cohort. The 24-25 CEIRR cohort is powered for NT50 comparisons by: - sex: male and female - age: binned into 3 groups: - 18-44 years of age - 45-55 years of age - 56-70 years of age

# Data Ingestion Instructions

## Ingested data should be present in 1 excel file across different worksheet (tabs)

1.  One tab contains serology data formatted by 'tidy' standards
2.  Another tab shoud contain a flat table with demographics including age and sex a,ong with patient ID and number for joining.

## How to Update Data:

1.  Navigate to the 2024-25 master serology excel spreadsheet
2.  Navigate to the "files" pane and click "upload"
3.  Navigate to the most rescent version of the spreadsheet and proceed to upload to the `source/` directory
4.  Rerun the data digest code block

Note: the cohort demographic data has been imported seperately and is located under `data/CEIRR_demographics.tsv`

# Import, clean and configure all values

1.  import serology
2.  impoty demographics
3.  join serology to demographics
4.  pivot long against nt50 and auc
5.  caluclate log2 and lo10 for auc
6.  set limit of detections for all values

```{r}
library(readxl)
library(tidyverse)
library(ggdist)
library(rstatix)
library(tidyplots)
library(plotly)

theme_set(theme_gray(base_family = "Helvetica"))

# Colors
timepoint_colors <- c("d0" = "#F2EBE2", "d28" = "#8F9779") 

virus_colors <- c(
  "A/Victoria/4897/2022" = "#ff7f0e", 
  "A/California/122/2022" = "#FCBA04", 
  "B/Austria/1359417/2021" = "#A50104", 
  "JH25260" = "#0c457d", 
  "JH25250" = "#6bd2db", 
  "JH25588" = "#0c457d", 
  "JH25610" = "#6bd2db",
  "JH24547" = "#FF5003", 
  "JH251192" = "#00A7E1"
)

vaccine_virus_short <- c(
  "A/Victoria/4897/2022" = "H1N1", 
  "A/California/122/2022" = "H3N2", 
  "B/Austria/1359417/2021" = "B/Victoria"
)

# Load and merge data
ceirr_serology <- read_xlsx("data/serology/NH2024-25_jhh_serology_titers.xlsx", sheet = "serology")
ceirr_demographics <- read_xlsx("data/serology/NH2024-25_jhh_serology_titers.xlsx", sheet = "demographics") %>%
  select(-sample_number)

# standarize all limits of detection to 10 - bring the 5's up to 10. 
ceirr <- left_join(ceirr_serology, ceirr_demographics, by = "study_id") |>
  mutate(
    nt50_d0 = if_else(!is.na(nt50_d0) & nt50_d0 < 10, 10, nt50_d0),
    nt50_d28 = if_else(!is.na(nt50_d28) & nt50_d28 < 10, 10, nt50_d28),
    auc_d0 = if_else(!is.na(auc_d0) & auc_d0 == 0, 1.25, auc_d0),
    auc_d28 = if_else(!is.na(auc_d28) & auc_d28 == 0, 1.25, auc_d28)
  )

# Tidy and transform long format
ceirr_long <- ceirr %>%
  pivot_longer(
    cols = c(nt50_d0, nt50_d28, auc_d0, auc_d28),
    names_to = c("measurement", "timepoint"),
    names_pattern = "(nt50|auc)_d(\\d+)",
    values_to = "value"
  ) %>%
  pivot_wider(names_from = measurement, values_from = value) %>%
  mutate(
    timepoint = paste0("d", timepoint),
    log2_nt50 = log2(nt50),
    log10_auc = log10(auc)
  )

# Calculate mean fold change
ceirr_fc_by_virus <- ceirr_long %>%
  filter(!is.na(nt50)) %>%
  select(study_id, virus, timepoint, nt50) %>%
  pivot_wider(names_from = timepoint, values_from = nt50) %>%
  group_by(virus) %>%
  mutate(fold_change = d28 / d0) %>%
  summarise(mean_fold_change = mean(fold_change, na.rm = TRUE), .groups = "drop") %>%
  mutate(label = paste0("x", round(mean_fold_change, 2)))

# Final combined dataset with fold change
ceirr_long_fc <- ceirr_long %>%
  left_join(ceirr_fc_by_virus, by = "virus")

# calculate patient fold change
ceirr_long_fc <- ceirr_long_fc %>%
  group_by(sample_number, virus) %>%
  mutate(
    patient_fold_change = {
      d0 <- nt50[timepoint == "d0"]
      d28 <- nt50[timepoint == "d28"]
      if (length(d0) == 1 && length(d28) == 1 && !is.na(d0) && !is.na(d28)) {
        d28 / d0
      } else {
        NA_real_
      }
    }
  ) %>%
  ungroup()

ceirr_long_fc <- ceirr_long_fc %>%
  group_by(sample_number, virus) %>%
  mutate(
    patient_fold_change = {
      d0 <- nt50[timepoint == "d0"]
      d28 <- nt50[timepoint == "d28"]
      if (length(d0) == 1 && length(d28) == 1 && !is.na(d0) && !is.na(d28)) {
        d28 / d0
      } else {
        NA_real_
      }
    },
    patient_fold_change_auc = {
      a0 <- auc[timepoint == "d0"]
      a28 <- auc[timepoint == "d28"]
      if (length(a0) == 1 && length(a28) == 1 && !is.na(a0) && !is.na(a28)) {
        a28 / a0
      } else {
        NA_real_
      }
    }
  ) %>%
  ungroup()

# define global limits of detection (lod or ld)
nt50ld <- log2(20)
aucld <- log10(2.5) #aucld <- log10(0.125) # based off a starting dilution of 1/20

# Calculate Number of patients with ≤LOD values for each vaccine virus
ceirr_sn_d0_nt50 = ceirr_long_fc |>
  filter(timepoint == "d0") |>
  group_by(virus) |>
  summarize(
    "Baseline Positive" = sum(nt50 > 10),
    "Baseline Negative" = sum(nt50 == 10)
  ) |>
    pivot_longer(cols = c("Baseline Negative", "Baseline Positive"), names_to = "Response", values_to = "Count")

ceirr_sn_d28_nt50 = ceirr_long_fc |>
  filter(timepoint == "d28") |>
  group_by(virus) |>
  summarize(
    "Baseline Positive" = sum(nt50 > 10),
    "Baseline Negative" = sum(nt50 == 10)
  ) |>
    pivot_longer(cols = c("Baseline Negative", "Baseline Positive"), names_to = "Response", values_to = "Count")

# auc seronegative
ceirr_sn_d0_auc = ceirr_long_fc |>
  filter(timepoint == "d0") |>
  group_by(virus) |>
  summarize(
    "Baseline Positive" = sum(auc > 2.5),
    "Baseline Negative" = sum(auc == 1.25)
  ) |>
    pivot_longer(cols = c("Baseline Negative", "Baseline Positive"), names_to = "Response", values_to = "Count")

ceirr_sn_d28_auc = ceirr_long_fc |>
  filter(timepoint == "d28") |>
  group_by(virus) |>
  summarize(
    "Baseline Positive" = sum(auc > 2.5),
    "Baseline Negative" = sum(auc == 1.25)
  ) |>
    pivot_longer(cols = c("Baseline Negative", "Baseline Positive"), names_to = "Response", values_to = "Count")

```

## IBV

### NT50 and AUC Pre- and Post- vax
```{r}
## Day 0 only
ceirr_long_fc |>
  filter(`timepoint` == 'd0', subtype == 'Victoria') |>
  tidyplot(x = virus, y = log2_nt50, color = virus) |>
  add_data_points_beeswarm(alpha= 0.8, shape = 1, size = 2) |>
  add_data_points_beeswarm(alpha= 0.6, shape = 21, color="black", size = 2) |> 
  adjust_colors(virus_colors) |>
  add_mean_dash(color = "black", linewidth = 0.85) |>
  add_sem_errorbar(color = "black") |>
  adjust_x_axis(rotate_labels = 45) |>
  adjust_y_axis(
    breaks = c(0, 2.5, 5, 7.5, 10, 12.5),
    limits = c(0, 14)) |>
  adjust_y_axis_title("$Log[2]~NT[50]$") |>
  add_reference_lines(y=log2(20), linetype = 2, linewidth = 0.25) |> # LOD = 5 here
  #add_annotation_text("LOD", x = 0.6, y = log2(20)) |> 
  add_title("Day 0") |>
  remove_legend() |>
  remove_x_axis_title() |>
  rename_x_axis_labels(new_names = c(
    "B/Austria/1359417/2021" = "Vaccine",
    "JH24547" = "23-24 C.5.1 ",
    "JH251192" = "24-25 C.3re" )) |>
  adjust_font(fontsize = 14) |>
  add_test_pvalue(method = "wilcox_test" , p.adjust.method = "bonferroni", hide.ns = TRUE) |> 
  save_plot("figures/ibv_d0_NT50.png", bg = "transparent")

## Day 28 Only
ceirr_long_fc %>% 
  filter(`timepoint` == 'd28', subtype == 'Victoria') %>% 
  tidyplot(x = virus, y = log2_nt50, color = virus) |>
  add_data_points_beeswarm(alpha= 0.8, shape = 2, size = 2) |>
  add_data_points_beeswarm(alpha= 0.6, shape = 24, color="black", size = 2) |> 
  adjust_colors(virus_colors) |>
  add_mean_dash(color = "black", linewidth = 0.85) |>
  add_sem_errorbar(color = "black") |>
  adjust_x_axis(rotate_labels = 45) |>
  adjust_y_axis(
    breaks = c(0, 2.5, 5, 7.5, 10, 12.5),
    limits = c(0, 14)) |>
  adjust_y_axis_title("$Log[2]~NT[50]$") |>
  add_reference_lines(y=log2(20), linetype = 2, linewidth = 0.25) |> # LOD = 5 here
  #add_annotation_text("LOD", x = 0.6, y = 3) |> 
  add_title("Day 28") |>
  remove_legend() |>
  remove_x_axis_title() |>
  rename_x_axis_labels(new_names = c(
    "B/Austria/1359417/2021" = "Vaccine",
    "JH24547" = "23-24 C.5.1 ",
    "JH251192" = "24-25 C.3re" )) |>
  adjust_font(fontsize = 14) |>
  add_test_pvalue(method = "wilcox_test" , p.adjust.method = "bonferroni", hide.ns = TRUE) |> 
  save_plot("figures/ibv_d28_NT50.png", bg = "transparent")

# AUC

## Day 0 Vaccines Only
ceirr_long_fc |>
  filter(`timepoint` == 'd0', subtype == 'Victoria', !is.na(log10_auc)) |> #drop any AUC rows that are NA 
  tidyplot(x = virus, y = log10_auc, color = virus, height= 50, width = 50) |>
  add_data_points_beeswarm(alpha= 0.8, shape = 1, size = 2) |>
  add_data_points_beeswarm(alpha= 0.6, shape = 21, color="black", size = 2) |> 
  adjust_colors(virus_colors) |>
  add_mean_dash(color = "black", linewidth = 0.85) |>
  add_sem_errorbar(color = "black") |>
  adjust_x_axis(rotate_labels = 45) |>
  adjust_y_axis(breaks = c(0, 1, 2, 3, 4, 5), limits = c(0, 5)) |>
  adjust_y_axis_title("$Log[10]~AUC$") |>
  add_reference_lines(y=log10(2.5), linetype = 2, linewidth = 0.25) |>
  #add_annotation_text("LOD", x = 0.6, y = 0.3) |> 
  add_title("Day 0") |>
  remove_legend() |>
  remove_x_axis_title() |>
  rename_x_axis_labels(new_names = c(
    "B/Austria/1359417/2021" = "Vaccine",
    "JH24547" = "23-24 C.5.1 ",
    "JH251192" = "24-25 C.3re" )) |>
  adjust_font(fontsize = 14) |>
  add_test_pvalue(method = "wilcox_test" , p.adjust.method = "bonferroni", hide.ns = TRUE) |> 
  save_plot("figures/ibv_d0_AUC.png", bg = "transparent")

## Day 28 Only
ceirr_long_fc |>
  filter(`timepoint` == 'd28', subtype == 'Victoria', !is.na(log10_auc)) |> #drop any AUC rows that are NA 
  tidyplot(x = virus, y = log10_auc, color = virus, height= 50, width = 50) |>
  add_data_points_beeswarm(alpha= 0.8, shape = 2, size = 2) |>
  add_data_points_beeswarm(alpha= 0.6, shape = 24, color="black", size = 2) |> 
  adjust_colors(virus_colors) |>
  add_mean_dash(color = "black", linewidth = 0.85) |>
  add_sem_errorbar(color = "black") |>
  adjust_x_axis(rotate_labels = 45) |>
  adjust_y_axis(breaks = c(0, 1, 2, 3, 4, 5), limits = c(0, 5)) |>
  adjust_y_axis_title("$Log[10]~AUC$") |>
  add_reference_lines(y=log10(2.5), linetype = 2, linewidth = 0.25) |>
  #add_annotation_text("LOD", x = 0.6, y = 0.3) |> 
  add_title("Day 28") |>
  remove_legend() |>
  remove_x_axis_title() |>
  rename_x_axis_labels(new_names = c(
    "B/Austria/1359417/2021" = "Vaccine",
    "JH24547" = "23-24 C.5.1 ",
    "JH251192" = "24-25 C.3re" )) |>
  adjust_font(fontsize = 14) |>
  add_test_pvalue(method = "wilcox_test" , p.adjust.method = "bonferroni", hide.ns = TRUE) |> 
  save_plot("figures/ibv_d28_AUC.png", bg = "transparent")

```

# NT50 Fold Change 
```{r}

# calculate patients ≥ 4 fold change by virus
# Vax = 19
# JH24547 = 18
# JH251192 = 1

ceirr_long_fc |>
  filter(subtype == "Victoria", timepoint == "d0") |>
  tidyplot(x = virus, y = patient_fold_change , color = virus) |>
  add_data_points_beeswarm(alpha= 1, shape = 0, size = 2) |>
  add_data_points_beeswarm(alpha= 0.6, shape = 22, color="black", size = 2) |> 
  add_annotation_line(y=7.9, yend=7.9, x=0.7, xend = 1.3) |> # Vaccine FC average manually enter
  add_annotation_line(y=6.2, yend=6.2, x=1.7, xend = 2.3) |> # JH24547 FC average manually enter
  add_annotation_line(y=1.2, yend=1.2, x=2.7, xend = 3.3) |> # JH251192 FC average manually enter
  # add fold change text 
  add_annotation_text("19/50", x = 1 , y = 0.45, fontsize = 10) |> # JH251192 FC average manually enter
  add_annotation_text("18/50", x = 2 , y = 0.45, fontsize = 10) |> # JH251192 FC average manually enter
  add_annotation_text("1/50", x = 3 , y = 0.45, fontsize = 10) |> # JH251192 FC average manually enter
  add_test_pvalue(method = "wilcox_test" , p.adjust.method = "bonferroni", hide.ns = TRUE) |> 
  adjust_colors(virus_colors) |>
  adjust_x_axis(rotate_labels = 45) |>
  adjust_y_axis_title("$NT[50]~Fold~Change$") |>
  adjust_y_axis(transform = "log2", breaks = c(0.5, 1, 4, 32, 512), limits = c(0.1, 2000)) |>
  add_reference_lines(y=4, linetype = 2, linewidth = 0.25) |> # SC
  #adjust_y_axis(transform = "log2", breaks = c(0.5, 1, 4, 32, 512), limits = c(0.1, 1500)) |>
  add_title("NT50\nFold Change") |>
  remove_legend() |>
  remove_x_axis_title() |>
  rename_x_axis_labels(new_names = c(
    "B/Austria/1359417/2021" = "Vaccine",
    "JH24547" = "23-24 C.5.1",
    "JH251192" = "24-25 C.3re")) |>
  adjust_font(fontsize = 14) |>
  save_plot("figures/ibv_fold_change.png", bg = "transparent") 
```


### NT50 pre-post ditribution - facet by ibv virus

```{r}
library(stats)

#filter for vic only
ibv_all_df = ceirr_long_fc %>% 
  filter(subtype == "Victoria")

# Reshape long to wide for paired test
ibv_wide <- ibv_all_df %>%
  select(study_id, virus, timepoint, log2_nt50) %>%
  pivot_wider(
    names_from = timepoint,
    values_from = log2_nt50
  ) %>%
  drop_na(d0, d28)  # ensure no missing values

ibv_wilcox_results <- ibv_wide %>%
  group_by(virus) %>%
  summarise(
    n = n(),
    p = wilcox.test(d28, d0, paired = TRUE)$p.value
  ) %>%
  mutate(
    p.adj = p.adjust(p, method = "BH"),
    p.adj.signif = case_when(
      p.adj <= 0.001 ~ "***",
      p.adj <= 0.01  ~ "**",
      p.adj <= 0.05  ~ "*",
      TRUE           ~ "ns"
    )
  )

# View ibv_wilcox_results
print(ibv_wilcox_results)

#join results
ibv_all_df <- ibv_all_df %>%
  left_join(ibv_wilcox_results, by = "virus")

# Create a summary dataframe with mean_fold_change for annotation
annotation_data <- ibv_all_df %>%
  group_by(virus) %>%
  summarise(
    mean_fold_change = first(mean_fold_change),
    p.adj = first(p.adj)
  )

# start marginal vic plot 
ibv_all_plot = ibv_all_df %>% 
  ggplot(aes(x = timepoint, y = log2_nt50, fill = timepoint)) +  
    stat_slab(color="black",
        aes(
            justification = ifelse(timepoint == "d0", 1.2, -.2),
            side = ifelse(timepoint == "d0", "left", "right")
        ),
        expand = FALSE,
        trim = TRUE,
        density = "unbounded",
        lwd = 0.5
    ) +
    geom_path(aes(group = study_id), alpha = 0.3, size = 0.5) + 
    geom_point(alpha = 0.3, size = 2, shape = 21) +
    facet_grid(~factor(virus)) + 
    ylab(bquote(Log[2]~NT[50])) +
    xlab("Timepoint") +
    theme_classic() +
    theme(
        legend.position = "none",
        axis.text.x = element_text(size = 14),  # Increase X-axis font size
        axis.text.y = element_text(size = 14), 
        axis.title = element_text(size = 14),
        strip.background = element_blank(),  # Remove boxes around facet titles
        strip.text = element_text(size = 14)  # Adjust facet title font size if needed
    ) + 
    scale_y_continuous(
        limits = c(2, 15),  # Set Y-axis limits
        breaks = c(2.5, 5, 7.5, 10, 12.5)  # Set Y-axis breaks
    ) +
    geom_hline(yintercept = log2(20), linetype = "dotted", color = "black") + # add limit of detection
    stat_summary(
        fun.data = "mean_cl_boot",
        linewidth = 1,
        size = 0.25,
        colour = "black",
        alpha = 0.9,
        position = position_nudge(x = ifelse(ibv_all_df$timepoint == "d0", -0.3, 0.3)) 
    )  + 
    geom_text(
        data = annotation_data,
        aes(x = 1.5, y = 13, label = paste0("p = ", signif(p.adj, 2))),
        inherit.aes = FALSE,
        size = 4
    ) +
    geom_text(
        data = annotation_data,
        aes(x = 1.5, y = 14, label = paste0("x", round(mean_fold_change, 1))),
        inherit.aes = FALSE,
        size = 4
    ) +
    scale_fill_manual(values = timepoint_colors)
  # Apply custom colors for timepoints

ggsave("figures/ibv_all_nt50.png", ibv_all_plot, height = 3, width = 8)

ibv_all_plot

```

#### Plot each virus individually 
Normalizing kernel across all viruses using `facet()` is diminishing the visual effect of the desigram of non C.3re viruses. Plot and save individually. We'll manually add fold change and p values in figma.
```{r}

plot_virus_nt50 <- function(virus_name) {
  # Filter the main data and annotations
  plot_data <- ibv_all_df |> filter(virus == virus_name)
  virus_annotations <- annotation_data |> filter(virus == virus_name)

  ggplot(plot_data, aes(x = timepoint, y = log2_nt50, fill = timepoint)) +  
    stat_slab(
      color = "black",
      aes(
        justification = ifelse(timepoint == "d0", 1.2, -.2),
        side = ifelse(timepoint == "d0", "left", "right")
      ),
      expand = FALSE,
      trim = TRUE,
      density = "unbounded",
      lwd = 0.5
    ) +
    geom_path(aes(group = study_id), alpha = 0.3, size = 0.5) + 
    geom_point(alpha = 0.3, size = 2, shape = 21) +
    facet_grid(~factor(virus)) + 
    ylab(bquote(Log[2]~NT[50])) +
    xlab("Timepoint") +
    theme_classic() +
    theme(
      legend.position = "none",
      axis.text.x = element_text(size = 14),
      axis.text.y = element_text(size = 14), 
      axis.title = element_text(size = 14),
      strip.background = element_blank(),
      strip.text = element_text(size = 14)
    ) + 
    scale_y_continuous(
      limits = c(2, 13),
      breaks = c(2.5, 5, 7.5, 10, 12.5)
    ) +
    geom_hline(yintercept = log2(20), linetype = "dotted", color = "black") +
    stat_summary(
      fun.data = "mean_cl_boot",
      linewidth = 1,
      size = 0.25,
      colour = "black",
      alpha = 0.9,
      position = position_nudge(x = ifelse(plot_data$timepoint == "d0", -0.3, 0.3)) 
    )  + 
    geom_text(
      data = virus_annotations,
      aes(x = 1.5, y = 11, label = paste0("p = ", signif(p.adj, 2))),
      inherit.aes = FALSE,
      size = 4
    ) +
    geom_text(
      data = virus_annotations,
      aes(x = 1.5, y = 12, label = paste0("x", round(mean_fold_change, 1))),
      inherit.aes = FALSE,
      size = 4
    ) +
    scale_fill_manual(values = timepoint_colors)
}

```

#### NT50 indivudual plots
```{r}

ibv_single_nt50_dist_vax <- plot_virus_nt50("B/Austria/1359417/2021")
ibv_single_nt50_dist_547 <- plot_virus_nt50("JH24547")
ibv_single_nt50_dist_1192 <- plot_virus_nt50("JH251192")

ggsave("figures/ibv_single_nt50_dist_vax.png",ibv_single_nt50_dist_vax, height = 3, width = 3)
ggsave("figures/ibv_single_nt50_dist_547.png",ibv_single_nt50_dist_547, height = 3, width = 3)
ggsave("figures/ibv_single_nt50_dist_1192.png",ibv_single_nt50_dist_1192, height = 3, width = 3)

```




# AUC pre-post ditribution - facet by ibv virus

# NT50

```{r}

library(stats)

#filter for vic only
ibv_all_df = ceirr_long_fc %>% 
  filter(subtype == "Victoria")

# Reshape long to wide for paired test
ibv_wide_auc <- ibv_all_df %>%
  select(study_id, virus, timepoint, log10_auc) %>%
  pivot_wider(
    names_from = timepoint,
    values_from = log10_auc
  ) %>%
  drop_na(d0, d28)  # ensure no missing values

ibv_wilcox_results_auc <- ibv_wide_auc %>%
  group_by(virus) %>%
  summarise(
    n = n(),
    p = wilcox.test(d28, d0, paired = TRUE)$p.value
  ) %>%
  mutate(
    p.adj = p.adjust(p, method = "BH"),
    p.adj.signif = case_when(
      p.adj <= 0.001 ~ "***",
      p.adj <= 0.01  ~ "**",
      p.adj <= 0.05  ~ "*",
      TRUE           ~ "ns"
    )
  )

#join results
ibv_all_df <- ibv_all_df %>%
  left_join(ibv_wilcox_results_auc, by = "virus")

# Create a summary dataframe with mean_fold_change for annotation
annotation_data_auc <- ibv_all_df %>%
  group_by(virus) %>%
  summarise(
    patient_fold_change_auc = mean(mean_fold_change, na.rm = TRUE),
    p.adj = min(p.adj, na.rm = TRUE)  # or appropriate summary stat
  )

ibv_all_plot_auc = ibv_all_df %>% 
  ggplot(aes(x = timepoint, y = log10_auc, fill = timepoint, shape = timepoint)) +  
    stat_slab(color = "black",
        aes(
            justification = ifelse(timepoint == "d0", 1.2, -.2),
            side = ifelse(timepoint == "d0", "left", "right")
        ),
        expand = FALSE,
        trim = TRUE,
        density = "unbounded",
        lwd = 0.5
    ) +
    geom_path(aes(group = study_id), alpha = 0.3, size = 0.5) + 
    geom_point(alpha = 0.3, size = 2, color = "black") +
    scale_shape_manual(values = c("d0" = 21, "d28" = 24)) +  # add custom shapes
    facet_grid(~factor(virus)) + 
    ylab(bquote(Log[10]~AUC)) +
    xlab("Timepoint") +
    theme_classic() +
    theme(
        legend.position = "none",
        axis.text.x = element_text(size = 14),  
        axis.text.y = element_text(size = 14), 
        axis.title = element_text(size = 14),
        strip.background = element_blank(), 
        strip.text = element_text(size = 14)
    ) + 
    geom_hline(yintercept = aucld, linetype = "dotted", color = "black") +
    stat_summary(
        fun.data = "mean_cl_boot",
        linewidth = 1,
        size = 0.25,
        colour = "black",
        alpha = 0.9,
        position = position_nudge(x = ifelse(ibv_all_df$timepoint == "d0", -0.3, 0.3)) 
    )  + 
    geom_text(
        data = annotation_data_auc,
        aes(x = 1.5, y = 4, label = paste0("p = ", signif(p.adj, 2))),
        inherit.aes = FALSE,
        size = 4
    ) +
    geom_text(
        data = annotation_data_auc,
        aes(x = 1.5, y = 4.4, label = paste0("x", round(patient_fold_change_auc, 1))),
        inherit.aes = FALSE,
        size = 4
    ) +
    scale_fill_manual(values = timepoint_colors)

ggsave("figures/ibv_all_auc.png", ibv_all_plot_auc, height = 3, width = 8)

ibv_all_plot_auc
```

# NT50 Fold Change IBV
```{r}

# calculate patients ≥ 4 fold change by virus
# Vax = 19
# JH24547 = 18
# JH251192 = 1

ceirr_long_fc |>
  filter(subtype == "Victoria", timepoint == "d0") |>
  tidyplot(x = virus, y = patient_fold_change , color = virus) |>
  add_data_points_beeswarm(alpha= 1, shape = 1, size = 2) |>
  add_data_points_beeswarm(alpha= 0.6, shape = 21, color="black", size = 2) |> 
  add_annotation_line(y=7.9, yend=7.9, x=0.7, xend = 1.3) |> # Vaccine FC average manually enter
  add_annotation_line(y=6.2, yend=6.2, x=1.7, xend = 2.3) |> # JH24547 FC average manually enter
  add_annotation_line(y=1.2, yend=1.2, x=2.7, xend = 3.3) |> # JH251192 FC average manually enter
  # add fold change text 
  add_annotation_text("19/50", x = 1 , y = 0.45, fontsize = 10) |> # JH251192 FC average manually enter
  add_annotation_text("18/50", x = 2 , y = 0.45, fontsize = 10) |> # JH251192 FC average manually enter
  add_annotation_text("1/50", x = 3 , y = 0.45, fontsize = 10) |> # JH251192 FC average manually enter
  #add_mean_value(fontsize = 11, vjust = -10.49) |> # for some reason, the mean is the log2 transformed mean
  #add_mean_dash(color = "black", linewidth = 0.85) |> # for some reason, the mean is the log2 transformed mean
  #add_sem_errorbar(color = "black") |> # for some reason, the mean is the log2 transformed mean
  add_test_pvalue(method = "wilcox_test" , p.adjust.method = "bonferroni", hide.ns = TRUE) |> 
  adjust_colors(virus_colors) |>
  adjust_x_axis(rotate_labels = 45) |>
  adjust_y_axis_title("$NT[50]~Fold~Change$") |>
  adjust_y_axis(transform = "log2", breaks = c(0.5, 1, 4, 32, 512), limits = c(0.1, 2000)) |>
  add_reference_lines(y=4, linetype = 2, linewidth = 0.25) |> # SC
  #adjust_y_axis(transform = "log2", breaks = c(0.5, 1, 4, 32, 512), limits = c(0.1, 1500)) |>
  add_title("NT50\nFold Change") |>
  remove_legend() |>
  remove_x_axis_title() |>
  rename_x_axis_labels(new_names = c(
    "B/Austria/1359417/2021" = "Vaccine",
    "JH24547" = "23-24 C.5.1",
    "JH251192" = "24-25 C.3re")) |>
  adjust_font(fontsize = 14) |>
  save_plot("figures/ibv_fold_change.png", bg = "transparent") 
```


# Table - AUC by sex and age AUC

```{r}
library(dplyr)
library(tidyr)
library(purrr)

# Function to calculate GMT or GM AUC
calc_gmt <- function(x) {
  x <- x[!is.na(x) & x > 0]
  if (length(x) < 2) return(NA)
  gmean <- exp(mean(log(x)))
  gsd <- exp(sd(log(x)))
  sprintf("%.0f (%.2f)", gmean, gsd)
}

# Set viruses and timepoints
viruses <- c("B/Austria/1359417/2021", "JH24547", "JH251192")
timepoints <- c("d0", "d28")
sex_groups <- c("Female", "Male")
age_groups <- c("18-44", "45-55", "56-70")

# Function to generate one table for a given measurement column: nt50 or auc
generate_summary_table <- function(data, value_col = "nt50") {
  summary_rows <- list()
  
  for (v in viruses) {
    for (tp in timepoints) {
      df_sub <- data %>%
        filter(virus == v, timepoint == tp)
      
      # By sex
      sex_gmts <- map_chr(sex_groups, ~{
        calc_gmt(df_sub[[value_col]][df_sub$sex == .x])
      }) %>% set_names(sex_groups)
      
      total_sex <- calc_gmt(df_sub[[value_col]])
      sex_p <- tryCatch({
        if (length(unique(df_sub$sex)) == 2) {
          wilcox.test(as.formula(paste(value_col, "~ sex")), data = df_sub)$p.value
        } else NA
      }, error = function(e) NA)
      
      # By age group
      age_gmts <- map_chr(age_groups, ~{
        calc_gmt(df_sub[[value_col]][df_sub$`JH-CEIRR age grouping` == .x])
      }) %>% set_names(age_groups)
      
      total_age <- calc_gmt(df_sub[[value_col]])
      age_p <- tryCatch({
        kruskal.test(as.formula(paste(value_col, "~ `JH-CEIRR age grouping`")), data = df_sub)$p.value
      }, error = function(e) NA)
      
      # Combine into one row
      row <- tibble(
        Variable = paste(v, tp, sep = " – "),
        Female = sex_gmts["Female"],
        Male = sex_gmts["Male"],
        Total = total_sex,
        p_value_sex = ifelse(is.na(sex_p), NA, formatC(sex_p, format = "f", digits = 3)),
        `18-44` = age_gmts["18-44"],
        `45-55` = age_gmts["45-55"],
        `56-70` = age_gmts["56-70"],
        Total_age = total_age,
        p_value_age = ifelse(is.na(age_p), NA, formatC(age_p, format = "f", digits = 3))
      )
      
      summary_rows[[length(summary_rows) + 1]] <- row
    }
  }
  
  bind_rows(summary_rows)
}

# Generate both tables
nt50_table <- generate_summary_table(ibv_all_df, value_col = "nt50")
auc_table <- generate_summary_table(ibv_all_df, value_col = "auc")

# Add a column to distinguish the two
nt50_table <- nt50_table %>% mutate(Measurement = "NT50")
auc_table  <- auc_table  %>% mutate(Measurement = "AUC")

# Optionally combine or print separately
combined_table <- bind_rows(nt50_table, auc_table) %>%
  select(Measurement, everything())

# View tables
print(nt50_table, n = Inf)
print(auc_table, n = Inf)
# or print(combined_table)

```

## Export table


```{r}
library(gt)
library(dplyr)
library(tidyr)
library(openxlsx)
library(stringr)

# --- Separate virus and timepoint for grouping ---
combined_table <- combined_table %>%
  separate(Variable, into = c("Virus", "Timepoint"), sep = " – ") %>%
  mutate(Timepoint = factor(Timepoint, levels = c("d0", "d28"))) %>%
  arrange(Timepoint, Virus, Measurement)

# --- Create nicely formatted GT table ---
gt_summary <- combined_table %>%
  gt(groupname_col = "Timepoint") %>%
  tab_header(
    title = "Neutralization Summary by Sex and Age Group",
    subtitle = "NT50 and AUC values across timepoints"
  ) %>%
  cols_label(
    Measurement = "Measurement",
    Virus = "Virus",
    Female = "Female",
    Male = "Male",
    Total = "Total",
    p_value_sex = "Sex p-value",
    `18-44` = "18–44",
    `45-55` = "45–55",
    `56-70` = "56–70",
    Total_age = "Total",
    p_value_age = "Age p-value"
  ) %>%
  fmt_missing(everything(), missing_text = "-")

# --- Preview the table ---
gt_summary

# --- Save GT table as HTML (optional for visual check) ---
gtsave(gt_summary, "neutralization_summary_table.html")

# --- Save the data as .xlsx ---
# Optionally: reorder for Excel
excel_out <- combined_table %>%
  select(Timepoint, Measurement, Virus, Female, Male, Total, p_value_sex,
         `18-44`, `45-55`, `56-70`, Total_age, p_value_age)

write.xlsx(excel_out, file = "neutralization_summary_table.xlsx", rowNames = FALSE)

```


